---
title: "Behavioral Delta Risk: Calibrated Attrition Scoring for Targeted Retention"
author: "Geoffrey Bruder"
date: "`r format(Sys.Date(), '%Y-%m-%d')`"
output:
  pdf_document:
    latex_engine: xelatex
    toc: true
    toc_depth: 3
editor_options:
  markdown:
    wrap: 72
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, fig.width = 7, fig.height = 5)
set.seed(2026)
```

### Overview

This is the finalized, stakeholder‑ready analysis for my Behavioral Delta Risk: Calibrated Attrition Scoring for Targeted Retention project. It adds: (1) ensemble ablation quantifying the incremental value of the behavioral‑change features (AUC and Brier deltas with bootstrap CIs), (2) permutation sanity check and repeated‑split stability checks, and (3) a concise A/B test design and ROI calculation with documented illustrative assumptions. 

---

# 1 Packages and data load

```{r packages}
packages <- c(
  "tidyverse","caret","kableExtra","pROC","gridExtra","glmnet",
  "randomForest","xgboost","vip","scales","recipes","rsample","yardstick","broom","fastshap","boot","pwr"
)
installed <- rownames(installed.packages())
for(p in packages) if(!(p %in% installed)) install.packages(p, dependencies = TRUE)
invisible(lapply(packages, library, character.only = TRUE))
```

```{r load-data}
data_path <- "BankChurners.csv"
if(!file.exists(data_path)) stop("BankChurners.csv not found in working directory.")
raw <- read.csv(data_path, stringsAsFactors = FALSE)

# Ensure anonymized PseudoID exists and is preserved (anonymized export id)
if(!"PseudoID" %in% names(raw)) raw <- raw %>% mutate(PseudoID = paste0("CUST", row_number()))

initial_cols <- names(raw)
message("Initial column count: ", length(initial_cols))
```

---

# 2 Targeted leakage audit and removal (preserve anonymized PseudoID)

```{r leakage-audit}
# Patterns that commonly indicate model outputs or leakage
leakage_patterns <- c("naive","classifier","predicted","risk","score","prob","probability","model")

# Identify columns matching patterns (case-insensitive)
mask_list <- lapply(leakage_patterns, function(p) grepl(p, names(raw), ignore.case = TRUE))
combined_mask <- if(length(mask_list) > 0) Reduce(`|`, mask_list) else rep(FALSE, length(names(raw)))
leakage_cols <- names(raw)[which(combined_mask)]

# Obvious real identifiers to remove from modeling (but keep anonymized PseudoID)
identifier_cols <- intersect(c("CLIENTNUM","Clientnum","ClientNum","clientnum","ID","Id","id"), names(raw))

# Build removal set but explicitly keep PseudoID
to_remove <- setdiff(unique(c(leakage_cols, identifier_cols)), "PseudoID")

# Create modeling-safe copy but keep PseudoID for final export
raw_clean <- raw %>% select(-any_of(to_remove))

message("Removed columns (leakage or identifiers): ", if(length(to_remove)>0) paste(to_remove, collapse = ", ") else "none")
message("Remaining column count: ", ncol(raw_clean))

# Audit table for review
audit_tbl <- tibble(
  Column = names(raw),
  Suspicious = names(raw) %in% leakage_cols,
  Identifier = names(raw) %in% identifier_cols
) %>%
  mutate(Reason = case_when(
    Suspicious ~ paste0("Matches pattern: ", paste(leakage_patterns[sapply(leakage_patterns, function(p) grepl(p, Column, ignore.case = TRUE))], collapse = ", ")),
    Identifier ~ "Identifier column",
    TRUE ~ NA_character_
  ))

if(length(to_remove) > 0){
  knitr::kable(tibble(Flagged = to_remove), caption = "Columns removed for leakage/identifier reasons")
} else {
  message("No leakage/identifier columns removed.")
}
```

---

# 3 Feature engineering and target

```{r preprocess}
# Standardize names
names(raw_clean) <- make.names(names(raw_clean))
data_full <- raw_clean

# Convert categorical variables to factors if present
possible_cat <- c("Attrition_Flag","Gender","Education_Level","Marital_Status","Income_Category","Card_Category")
cat_vars <- intersect(possible_cat, names(data_full))
if(length(cat_vars) > 0){
  data_full[cat_vars] <- lapply(data_full[cat_vars], function(x) ifelse(is.na(x) | x=="", "Unknown", x))
  data_full[cat_vars] <- lapply(data_full[cat_vars], factor)
}

# Numeric coercion for expected numeric columns
possible_num <- c(
  "Customer_Age","Dependent_count","Months_on_book","Total_Relationship_Count",
  "Months_Inactive_12_mon","Contacts_Count_12_mon","Credit_Limit","Total_Revolving_Bal",
  "Avg_Open_To_Buy","Total_Amt_Chng_Q4_Q1","Total_Trans_Amt","Total_Trans_Ct",
  "Total_Ct_Chng_Q4_Q1","Avg_Utilization_Ratio"
)
num_vars <- intersect(possible_num, names(data_full))
data_full[num_vars] <- lapply(data_full[num_vars], function(x) as.numeric(as.character(x)))

# Behavioral-change features (novelty)
data_full <- data_full %>%
  mutate(
    Amt_Chg_Q4Q1 = ifelse("Total_Amt_Chng_Q4_Q1" %in% names(.), Total_Amt_Chng_Q4_Q1, NA_real_),
    Ct_Chg_Q4Q1  = ifelse("Total_Ct_Chng_Q4_Q1" %in% names(.), Total_Ct_Chng_Q4_Q1, NA_real_),
    Large_Amt_Drop = ifelse(!is.na(Amt_Chg_Q4Q1) & Amt_Chg_Q4Q1 < 0.7, 1, 0),
    Large_Ct_Drop  = ifelse(!is.na(Ct_Chg_Q4Q1) & Ct_Chg_Q4Q1 < 0.7, 1, 0),
    Recent_Contact_Reduction = ifelse(!is.na(Contacts_Count_12_mon) & Contacts_Count_12_mon <= 1, 1, 0)
  )

# Binary target
if("Attrition_Flag" %in% names(data_full)){
  data_full <- data_full %>%
    mutate(target_attrit = case_when(
      grepl("Attrited", Attrition_Flag, ignore.case = TRUE) ~ 1,
      grepl("Existing", Attrition_Flag, ignore.case = TRUE) ~ 0,
      TRUE ~ NA_real_
    ))
} else {
  data_full$target_attrit <- NA_real_
}

# Keep a copy for final scoring
data_for_scoring <- data_full

# Modeling subset
model_data <- data_full %>% filter(!is.na(target_attrit))
message("Rows for modeling: ", nrow(model_data))
```

---

# 4 Reproducible pipeline, train/test split, and predictor audit

```{r split-recipe-audit}
set.seed(2026)
split <- initial_split(model_data, prop = 0.7, strata = "target_attrit")
train <- training(split)
test  <- testing(split)

# Exclude identifiers and target (ensure CLIENTNUM not used)
exclude_cols <- c("PseudoID","Attrition_Flag","target_attrit","CLIENTNUM","Clientnum","clientnum")
all_predictors <- setdiff(names(train), exclude_cols)

# Build recipe (PseudoID excluded from predictors by exclude_cols)
rec <- recipe(target_attrit ~ ., data = train %>% select(all_of(c("target_attrit", all_predictors)))) %>%
  step_impute_median(all_numeric(), -all_outcomes()) %>%
  step_unknown(all_nominal(), -all_outcomes()) %>%
  step_other(all_nominal(), -all_outcomes(), threshold = 0.01) %>%
  step_dummy(all_nominal(), -all_outcomes()) %>%
  step_zv(all_predictors()) %>%
  step_normalize(all_numeric(), -all_outcomes())

prep_rec <- prep(rec, training = train)
train_baked <- bake(prep_rec, new_data = train)
test_baked  <- bake(prep_rec, new_data = test)

# caret-friendly factors
train_baked <- train_baked %>% mutate(target = factor(ifelse(target_attrit==1,"Yes","No")))
test_baked  <- test_baked  %>% mutate(target = factor(ifelse(target_attrit==1,"Yes","No")))

# x/y sets
x_train <- train_baked %>% select(-target_attrit, -target)
y_train <- train_baked$target
x_test  <- test_baked %>% select(-target_attrit, -target)
y_test  <- test_baked$target

# Ensure numeric and data.frame
x_train <- x_train %>% mutate(across(everything(), ~ as.numeric(as.character(.))))
x_test  <- x_test  %>% mutate(across(everything(), ~ as.numeric(as.character(.))))
x_train <- as.data.frame(x_train)
x_test  <- as.data.frame(x_test)

# Impute remaining NA/Inf with medians
na_cols <- colSums(is.na(x_train) | is.infinite(as.matrix(x_train)))
if(any(na_cols > 0)){
  for(col in names(na_cols)[na_cols > 0]){
    med <- median(x_train[[col]], na.rm = TRUE)
    x_train[[col]][is.na(x_train[[col]]) | is.infinite(x_train[[col]])] <- med
    x_test[[col]][is.na(x_test[[col]]) | is.infinite(x_test[[col]])] <- med
  }
}

message("Prepared x_train: ", nrow(x_train), " x ", ncol(x_train))

# Predictor audit
predictor_list <- names(x_train)
predictor_audit <- tibble(Predictor = predictor_list) %>%
  mutate(Suspicious = grepl(paste(leakage_patterns, collapse = "|"), Predictor, ignore.case = TRUE))
knitr::kable(predictor_audit, caption = "Predictors used in modeling and whether they match leakage patterns")
message("Number of predictors used: ", length(predictor_list))
```

---

# 5 Baseline and enhanced logistic (unchanged functionality)

```{r baseline-enhanced}
# Baseline static predictors
static_preds <- intersect(c("Customer_Age","Gender","Credit_Limit","Total_Trans_Amt","Total_Trans_Ct",
                            "Avg_Utilization_Ratio","Total_Relationship_Count","Months_on_book"), names(train))
static_preds <- static_preds[static_preds != ""]

rec_base <- recipe(as.formula(paste("target_attrit ~", paste(static_preds, collapse = " + "))), data = train) %>%
  step_impute_median(all_numeric(), -all_outcomes()) %>%
  step_unknown(all_nominal(), -all_outcomes()) %>%
  step_dummy(all_nominal(), -all_outcomes()) %>%
  step_normalize(all_numeric(), -all_outcomes())

train_base <- bake(prep(rec_base, training = train), new_data = train)
test_base  <- bake(prep(rec_base, training = train), new_data = test)

glm_base <- glm(target_attrit ~ ., data = train_base, family = binomial)
pred_base_prob <- predict(glm_base, newdata = test_base, type = "response")
roc_base <- pROC::roc(test_base$target_attrit, pred_base_prob, quiet = TRUE)
auc_base <- pROC::auc(roc_base)

# Enhanced logistic: all baked predictors
enh_predictors <- names(train_baked)[!names(train_baked) %in% c("target_attrit","target")]
train_enh_df <- train_baked %>% select(all_of(c("target_attrit", enh_predictors)))
test_enh_df  <- test_baked  %>% select(all_of(c("target_attrit", enh_predictors)))

glm_enh <- glm(target_attrit ~ ., data = train_enh_df, family = binomial)
pred_enh_prob <- predict(glm_enh, newdata = test_enh_df, type = "response")
roc_enh <- pROC::roc(test_enh_df$target_attrit, pred_enh_prob, quiet = TRUE)
auc_enh <- pROC::auc(roc_enh)

# Brier scores
brier_base <- mean((pred_base_prob - test_base$target_attrit)^2, na.rm = TRUE)
brier_enh  <- mean((pred_enh_prob - test_enh_df$target_attrit)^2, na.rm = TRUE)

knitr::kable(
  tibble(Model = c("Logit Baseline","Logit Enhanced"),
         AUC = c(as.numeric(auc_base), as.numeric(auc_enh)),
         Brier = c(brier_base, brier_enh)),
  caption = "Baseline vs Enhanced Logistic: AUC and Brier"
)
```

---

# 6 Random Forest: tuning, repeated CV stability, and importance

```{r rf-stability}
set.seed(2026)
ctrl_rep <- trainControl(method = "repeatedcv", number = 5, repeats = 5,
                         classProbs = TRUE, summaryFunction = twoClassSummary,
                         savePredictions = "final", verboseIter = FALSE, allowParallel = FALSE)

rf_grid <- expand.grid(mtry = c(3,5,7))
rf_model <- train(x = x_train, y = y_train,
                  method = "rf",
                  metric = "ROC",
                  trControl = ctrl_rep,
                  tuneGrid = rf_grid,
                  ntree = 500)

# Resample AUCs
rf_resamples <- rf_model$resample$ROC
rf_mean_auc <- mean(rf_resamples, na.rm = TRUE)
rf_sd_auc   <- sd(rf_resamples, na.rm = TRUE)

# Test set predictions
rf_pred_prob <- predict(rf_model, newdata = x_test, type = "prob")[, "Yes"]
roc_rf <- pROC::roc(ifelse(y_test=="Yes",1,0), rf_pred_prob, quiet = TRUE)
auc_rf <- pROC::auc(roc_rf)
brier_rf <- mean((rf_pred_prob - ifelse(y_test=="Yes",1,0))^2, na.rm = TRUE)

message("Random Forest test AUC: ", round(as.numeric(auc_rf),4),
        " | CV mean AUC: ", round(rf_mean_auc,4), " ± ", round(rf_sd_auc,4))

# Variable importance (top 20)
rf_varimp <- varImp(rf_model)$importance %>% rownames_to_column("Feature") %>% arrange(desc(Overall))
top20_rf <- head(rf_varimp, 20)
knitr::kable(top20_rf, caption = "Random Forest variable importance (top 20)")

# VIP plot (best-effort)
try({ vip::vip(rf_model$finalModel, num_features = 20) }, silent = TRUE)
```

---

# 7 XGBoost: caret repeated CV stability and direct xgboost fallback

```{r xgb-stability-and-final}
set.seed(2026)
ctrl_xgb <- trainControl(method = "repeatedcv", number = 5, repeats = 5,
                         classProbs = TRUE, summaryFunction = twoClassSummary,
                         savePredictions = "final", verboseIter = FALSE, allowParallel = FALSE)

xgb_grid_caret <- expand.grid(nrounds = c(50,100),
                              max_depth = c(3,6),
                              eta = c(0.05,0.1),
                              gamma = 0,
                              colsample_bytree = c(0.7,0.8),
                              min_child_weight = c(1,3),
                              subsample = c(0.7,0.8))

xgb_caret <- tryCatch({
  train(x = x_train, y = y_train,
        method = "xgbTree",
        metric = "ROC",
        trControl = ctrl_xgb,
        tuneGrid = xgb_grid_caret,
        verbose = FALSE)
}, error = function(e){
  warning("caret xgbTree failed: ", e$message)
  NULL
})

# Fallback to direct xgboost CV grid if caret failed
if(is.null(xgb_caret)){
  message("Falling back to direct xgboost CV grid.")
  param_grid <- expand.grid(max_depth = c(3,6),
                            eta = c(0.05,0.1),
                            subsample = c(0.7,0.8),
                            colsample_bytree = c(0.7,0.8),
                            min_child_weight = c(1,3),
                            nrounds = c(50,100))
  best_result <- NULL; best_auc <- -Inf
  dtrain <- xgboost::xgb.DMatrix(data = as.matrix(x_train), label = ifelse(y_train=="Yes",1L,0L))
  for(i in seq_len(nrow(param_grid))){
    params <- list(objective = "binary:logistic", eval_metric = "auc",
                   max_depth = param_grid$max_depth[i],
                   eta = param_grid$eta[i],
                   subsample = param_grid$subsample[i],
                   colsample_bytree = param_grid$colsample_bytree[i],
                   min_child_weight = param_grid$min_child_weight[i],
                   verbosity = 0)
    cv <- tryCatch({
      xgboost::xgb.cv(params = params, data = dtrain, nrounds = param_grid$nrounds[i],
                      nfold = 5, early_stopping_rounds = 20, verbose = FALSE, maximize = TRUE)
    }, error = function(e) NULL)
    if(!is.null(cv)){
      eval_log <- as.data.frame(cv$evaluation_log)
      auc_col <- grep("test_auc", names(eval_log), value = TRUE)[1]
      best_iter <- if(!is.null(cv$best_iteration) && cv$best_iteration > 0) as.integer(cv$best_iteration) else which.max(eval_log[[auc_col]])
      if(length(best_iter) == 0 || !is.finite(best_iter) || best_iter < 1) next
      best_score <- max(eval_log[[auc_col]], na.rm = TRUE)
      if(best_score > best_auc){
        best_auc <- best_score
        best_result <- list(params = params, nrounds = as.integer(best_iter), cv = cv)
      }
    }
  }
  if(is.null(best_result)) stop("Direct xgboost CV failed for all parameter combinations.")
  nrounds_final <- best_result$nrounds
  dtrain_final <- xgboost::xgb.DMatrix(data = as.matrix(x_train), label = ifelse(y_train=="Yes",1L,0L))
  xgb_final <- xgboost::xgb.train(params = best_result$params, data = dtrain_final, nrounds = nrounds_final, verbose = 0, watchlist = list(train = dtrain_final), maximize = TRUE)
  xgb_caret <- list(bestTune = data.frame(nrounds = nrounds_final,
                                          max_depth = best_result$params$max_depth,
                                          eta = best_result$params$eta,
                                          gamma = 0,
                                          colsample_bytree = best_result$params$colsample_bytree,
                                          min_child_weight = best_result$params$min_child_weight,
                                          subsample = best_result$params$subsample),
                    resample = data.frame(ROC = best_auc))
  class(xgb_caret) <- "fallback_xgb_caret"
} else {
  # caret succeeded: build direct xgboost final model using caret bestTune
  best_caret_params <- xgb_caret$bestTune
  params_direct <- list(
    objective = "binary:logistic",
    eval_metric = "auc",
    max_depth = best_caret_params$max_depth,
    eta = best_caret_params$eta,
    subsample = best_caret_params$subsample,
    colsample_bytree = best_caret_params$colsample_bytree,
    min_child_weight = best_caret_params$min_child_weight,
    verbosity = 0
  )
  dtrain <- xgboost::xgb.DMatrix(data = as.matrix(x_train), label = ifelse(y_train=="Yes",1L,0L))
  cv_direct <- tryCatch({
    xgboost::xgb.cv(params = params_direct, data = dtrain, nrounds = best_caret_params$nrounds,
                    nfold = 5, early_stopping_rounds = 20, verbose = FALSE, maximize = TRUE)
  }, error = function(e) NULL)
  nrounds_final <- if(!is.null(cv_direct) && !is.null(cv_direct$best_iteration) && cv_direct$best_iteration > 0) as.integer(cv_direct$best_iteration) else best_caret_params$nrounds
  xgb_final <- xgboost::xgb.train(params = params_direct, data = dtrain, nrounds = nrounds_final, verbose = 0, watchlist = list(train = dtrain), maximize = TRUE)
}

# Test predictions
dtest <- xgboost::xgb.DMatrix(data = as.matrix(x_test))
xgb_pred_prob <- predict(xgb_final, newdata = dtest)
roc_xgb <- pROC::roc(ifelse(y_test=="Yes",1,0), xgb_pred_prob, quiet = TRUE)
auc_xgb <- pROC::auc(roc_xgb)
brier_xgb <- mean((xgb_pred_prob - ifelse(y_test=="Yes",1,0))^2, na.rm = TRUE)

# caret-like stability metrics if available
xgb_resamples <- if(!is.null(xgb_caret$resample$ROC)) xgb_caret$resample$ROC else c(auc_xgb)
xgb_mean_auc <- mean(xgb_resamples, na.rm = TRUE)
xgb_sd_auc   <- sd(xgb_resamples, na.rm = TRUE)

message("XGBoost test AUC: ", round(as.numeric(auc_xgb),4),
        " | CV mean AUC: ", round(xgb_mean_auc,4), " ± ", round(xgb_sd_auc,4))

# XGBoost variable importance (top 20)
imp_xgb <- tryCatch(xgboost::xgb.importance(model = xgb_final), error = function(e) NULL)
if(!is.null(imp_xgb)){
  top20_xgb <- head(imp_xgb, 20)
  knitr::kable(top20_xgb, caption = "XGBoost variable importance (top 20)")
  try({ vip::vip(xgb_final, num_features = 20) }, silent = TRUE)
} else {
  message("xgb.importance failed; skipping importance table.")
}

# Approximate SHAP (fastshap) for top features (best-effort)
if(!is.null(imp_xgb) && nrow(imp_xgb) > 0){
  top_features <- head(imp_xgb$Feature, 20)
  pred_fun <- function(object, newdata) {
    as.numeric(predict(object, newdata = xgboost::xgb.DMatrix(as.matrix(newdata))))
  }
  set.seed(2026)
  sample_idx <- sample(seq_len(nrow(x_train)), size = min(500, nrow(x_train)))
  X_sample <- as.data.frame(x_train[sample_idx, intersect(top_features, names(x_train)), drop = FALSE])
  shap_vals <- tryCatch({
    fastshap::explain(xgb_final, X = as.data.frame(x_train[, intersect(top_features, names(x_train)), drop = FALSE]), pred_wrapper = pred_fun, nsim = 50)
  }, error = function(e) NULL)
  if(!is.null(shap_vals)){
    shap_mean <- as.data.frame(abs(shap_vals)) %>% summarise_all(mean) %>% pivot_longer(everything(), names_to = "Feature", values_to = "MeanAbsSHAP") %>% arrange(desc(MeanAbsSHAP))
    knitr::kable(head(shap_mean, 20), caption = "Approximate SHAP mean absolute values (top features)")
  }
}
```

---

# 8 Model comparison and ROC plot

```{r compare}
auc_table <- tibble(
  Model = c("Logit Baseline","Logit Enhanced","Random Forest","XGBoost"),
  AUC = c(as.numeric(auc_base), as.numeric(auc_enh), as.numeric(auc_rf), as.numeric(auc_xgb)),
  Brier = c(brier_base, brier_enh, brier_rf, brier_xgb)
)
knitr::kable(auc_table, caption = "Model comparison on test set (AUC and Brier)")

# ROC plot
plot(roc_base, col = "black", main = "ROC curves (test set)")
plot(roc_enh, add = TRUE, col = "blue")
plot(roc_rf, add = TRUE, col = "darkgreen")
plot(roc_xgb, add = TRUE, col = "red")
legend("bottomright", legend = auc_table$Model, col = c("black","blue","darkgreen","red"), lwd = 2)
```

---

# 9 Recalibration (Platt scaling) and calibration plot

```{r recalibration}
# Select best model by test AUC
aucs <- c(as.numeric(auc_base), as.numeric(auc_enh), as.numeric(auc_rf), as.numeric(auc_xgb))
names(aucs) <- c("base","enh","rf","xgb")
best_model_name <- names(aucs)[which.max(aucs)]
best_model_name

# Raw probabilities on test set for selected model
raw_probs <- switch(best_model_name,
                    base = pred_base_prob,
                    enh  = pred_enh_prob,
                    rf   = rf_pred_prob,
                    xgb  = xgb_pred_prob)

# Platt scaling
eps <- 1e-6
p_bounded <- pmin(pmax(raw_probs, eps), 1 - eps)
logit_raw <- log(p_bounded / (1 - p_bounded))
platt_df <- tibble(logit_raw = logit_raw, y = test_baked$target_attrit)
platt_model <- glm(y ~ logit_raw, data = platt_df, family = binomial)

# Apply Platt scaling to test set
platt_pred <- predict(platt_model, newdata = tibble(logit_raw = logit_raw), type = "response")

# Metrics before/after
brier_before <- mean((raw_probs - test_baked$target_attrit)^2, na.rm = TRUE)
brier_after  <- mean((platt_pred - test_baked$target_attrit)^2, na.rm = TRUE)
roc_before <- pROC::roc(test_baked$target_attrit, raw_probs, quiet = TRUE)
roc_after  <- pROC::roc(test_baked$target_attrit, platt_pred, quiet = TRUE)
auc_before <- pROC::auc(roc_before)
auc_after  <- pROC::auc(roc_after)

knitr::kable(
  tibble(Metric = c("Selected model","AUC before","AUC after","Brier before","Brier after"),
         Value = c(best_model_name, round(as.numeric(auc_before),4), round(as.numeric(auc_after),4), round(brier_before,6), round(brier_after,6))),
  caption = "Recalibration (Platt scaling) results on test set"
)

# Reliability diagram (binned)
cal_df <- tibble(raw = raw_probs, platt = platt_pred, truth = test_baked$target_attrit)
cal_plot <- cal_df %>%
  mutate(bin = ntile(platt, 10)) %>%
  group_by(bin) %>%
  summarize(mean_prob = mean(platt), obs = mean(truth), n = n(), .groups = "drop")
ggplot(cal_plot, aes(x = mean_prob, y = obs)) +
  geom_point(size = 2) + geom_line() + geom_abline(slope = 1, intercept = 0, linetype = "dashed") +
  labs(x = "Mean predicted probability (Platt)", y = "Observed event rate", title = "Reliability diagram for Platt calibrated selected model")
```

---

# 10 Decision threshold table

```{r decision-thresholds}
# Use calibrated probabilities for thresholding if calibration improved Brier
use_probs <- if(brier_after <= brier_before) platt_pred else raw_probs
truth <- test_baked$target_attrit

thresholds <- c(0.01, 0.02, 0.05, 0.10, 0.20)
threshold_table <- map_dfr(thresholds, function(t){
  preds_bin <- as.integer(use_probs >= t)
  tp <- sum(preds_bin == 1 & truth == 1)
  fp <- sum(preds_bin == 1 & truth == 0)
  tn <- sum(preds_bin == 0 & truth == 0)
  fn <- sum(preds_bin == 0 & truth == 1)
  precision <- if((tp+fp) == 0) NA_real_ else tp / (tp + fp)
  recall <- if((tp+fn) == 0) NA_real_ else tp / (tp + fn)
  tibble(threshold = t, TP = tp, FP = fp, TN = tn, FN = fn, Precision = precision, Recall = recall)
})

knitr::kable(threshold_table, digits = 4, caption = "Decision thresholds: expected TP/FP and precision/recall on test set")
```

---

# 11 Final scoring with recalibration and sample export

```{r final-scoring}
# Bake full dataset
full_baked <- bake(prep_rec, new_data = data_for_scoring)

# Compute final raw probabilities on full dataset using selected model
if(best_model_name == "rf"){
  final_x <- full_baked %>% select(any_of(names(x_train)))
  raw_full_probs <- predict(rf_model, newdata = final_x, type = "prob")[, "Yes"]
} else if(best_model_name == "xgb"){
  final_x <- full_baked %>% select(any_of(names(x_train)))
  final_matrix <- as.matrix(final_x %>% mutate(across(everything(), ~ as.numeric(as.character(.)))))
  dfinal <- xgboost::xgb.DMatrix(data = final_matrix)
  raw_full_probs <- predict(xgb_final, newdata = dfinal)
} else if(best_model_name == "enh"){
  final_baked_enh <- bake(prep(rec, training = train), new_data = data_for_scoring)
  raw_full_probs <- predict(glm_enh, newdata = final_baked_enh, type = "response")
} else {
  final_baked_base <- bake(prep(rec_base, training = train), new_data = data_for_scoring)
  raw_full_probs <- predict(glm_base, newdata = final_baked_base, type = "response")
}

# Apply Platt scaling to full dataset raw probabilities
p_bounded_full <- pmin(pmax(raw_full_probs, eps), 1 - eps)
logit_full <- log(p_bounded_full / (1 - p_bounded_full))
calibrated_full <- predict(platt_model, newdata = tibble(logit_raw = logit_full), type = "response")

data_for_scoring$Predicted_Risk_Raw <- raw_full_probs
data_for_scoring$Predicted_Risk_Calibrated <- calibrated_full

# Safe selection for sample final predictions
cols_wanted <- c("PseudoID","Predicted_Risk_Raw","Predicted_Risk_Calibrated","Attrition_Flag")
cols_present <- intersect(cols_wanted, names(data_for_scoring))
knitr::kable(head(data_for_scoring %>% select(any_of(cols_present)), 10),
             caption = "Sample final predictions (raw and calibrated)")
```

---

# 12 Ablation study (behavioral features): ensembles with and without behavioral-change features

```{r ablation-setup}
# Identify behavioral feature names used in recipe (the engineered ones)
behavioral_feats <- c("Amt_Chg_Q4Q1","Ct_Chg_Q4Q1","Large_Amt_Drop","Large_Ct_Drop","Recent_Contact_Reduction")
behavioral_present <- intersect(behavioral_feats, names(train_baked))
behavioral_present
```

```{r ablation-models}
# Build two versions of training/test baked data: with and without behavioral features
non_behavioral_preds <- setdiff(names(x_train), behavioral_present)
x_train_no_beh <- x_train %>% select(any_of(non_behavioral_preds))
x_test_no_beh  <- x_test  %>% select(any_of(non_behavioral_preds))

# Train RF and XGB on both sets (keep same tuning choices; small grids to keep runtime reasonable)
set.seed(2026)
# RF with behavioral
rf_with <- train(x = x_train, y = y_train, method = "rf", metric = "ROC", trControl = ctrl_rep, tuneGrid = rf_grid, ntree = 500)
# RF without behavioral
rf_no  <- train(x = x_train_no_beh, y = y_train, method = "rf", metric = "ROC", trControl = ctrl_rep, tuneGrid = rf_grid, ntree = 500)

# XGB direct training (use previously selected params if available)
# For speed, use caret fallback if available; otherwise use direct xgboost params from best_result
if(!is.null(xgb_caret) && !inherits(xgb_caret, "fallback_xgb_caret")){
  xgb_with <- xgb_caret # caret object already trained on x_train
  # retrain caret on no-beh features by re-running train with same grid but reduced predictors
  xgb_no <- tryCatch({
    train(x = x_train_no_beh, y = y_train, method = "xgbTree", metric = "ROC", trControl = ctrl_xgb, tuneGrid = xgb_grid_caret, verbose = FALSE)
  }, error = function(e) NULL)
} else {
  # direct xgboost training using best_result$params and nrounds_final
  params_use <- if(exists("best_result")) best_result$params else list(objective="binary:logistic", eval_metric="auc", max_depth=3, eta=0.1, verbosity=0)
  nrounds_use <- if(exists("nrounds_final")) nrounds_final else 50
  dtrain_with <- xgboost::xgb.DMatrix(data = as.matrix(x_train), label = ifelse(y_train=="Yes",1L,0L))
  xgb_with <- xgboost::xgb.train(params = params_use, data = dtrain_with, nrounds = nrounds_use, verbose = 0)
  dtrain_no <- xgboost::xgb.DMatrix(data = as.matrix(x_train_no_beh), label = ifelse(y_train=="Yes",1L,0L))
  xgb_no <- xgboost::xgb.train(params = params_use, data = dtrain_no, nrounds = nrounds_use, verbose = 0)
}
```

```{r ablation-eval}
# Evaluate RF models on test set
rf_with_prob <- predict(rf_with, newdata = x_test, type = "prob")[, "Yes"]
rf_no_prob   <- predict(rf_no,  newdata = x_test_no_beh, type = "prob")[, "Yes"]

rf_with_auc <- as.numeric(pROC::auc(pROC::roc(ifelse(y_test=="Yes",1,0), rf_with_prob)))
rf_no_auc   <- as.numeric(pROC::auc(pROC::roc(ifelse(y_test=="Yes",1,0), rf_no_prob)))
rf_with_brier <- mean((rf_with_prob - ifelse(y_test=="Yes",1,0))^2, na.rm = TRUE)
rf_no_brier   <- mean((rf_no_prob - ifelse(y_test=="Yes",1,0))^2, na.rm = TRUE)

# Evaluate XGB models on test set (handle caret vs direct)
if(!is.null(xgb_caret) && !inherits(xgb_caret, "fallback_xgb_caret")){
  xgb_with_prob <- predict(xgb_caret, newdata = x_test, type = "prob")[, "Yes"]
  if(!is.null(xgb_no)){
    xgb_no_prob <- predict(xgb_no, newdata = x_test_no_beh, type = "prob")[, "Yes"]
  } else {
    xgb_no_prob <- rep(NA_real_, nrow(x_test_no_beh))
  }
} else {
  xgb_with_prob <- predict(xgb_with, newdata = xgboost::xgb.DMatrix(data = as.matrix(x_test)))
  xgb_no_prob   <- predict(xgb_no,  newdata = xgboost::xgb.DMatrix(data = as.matrix(x_test_no_beh)))
}

xgb_with_auc <- as.numeric(pROC::auc(pROC::roc(ifelse(y_test=="Yes",1,0), xgb_with_prob)))
xgb_no_auc   <- as.numeric(pROC::auc(pROC::roc(ifelse(y_test=="Yes",1,0), xgb_no_prob)))
xgb_with_brier <- mean((xgb_with_prob - ifelse(y_test=="Yes",1,0))^2, na.rm = TRUE)
xgb_no_brier   <- mean((xgb_no_prob - ifelse(y_test=="Yes",1,0))^2, na.rm = TRUE)

ablation_tbl <- tibble(
  Model = c("RF with behavioral","RF without behavioral","XGB with behavioral","XGB without behavioral"),
  AUC = c(rf_with_auc, rf_no_auc, xgb_with_auc, xgb_no_auc),
  Brier = c(rf_with_brier, rf_no_brier, xgb_with_brier, xgb_no_brier)
)
knitr::kable(ablation_tbl, digits = 4, caption = "Ablation: ensembles with vs without behavioral features (test set)")
```

---

# 13 Bootstrap CIs for ablation deltas (AUC and Brier)

```{r ablation-bootstrap}
# Function to compute AUC and Brier for a given resample index vector
compute_metrics <- function(idx, x_full_with, x_full_no, y_vec, model_with, model_no, model_type = c("rf","xgb")){
  xw <- x_full_with[idx, , drop = FALSE]
  xn <- x_full_no[idx, , drop = FALSE]
  y  <- y_vec[idx]
  if(model_type[1] == "rf"){
    p_with <- predict(model_with, newdata = xw, type = "prob")[, "Yes"]
    p_no   <- predict(model_no,  newdata = xn, type = "prob")[, "Yes"]
  } else {
    p_with <- predict(model_with, newdata = xgboost::xgb.DMatrix(as.matrix(xw)))
    p_no   <- predict(model_no,  newdata = xgboost::xgb.DMatrix(as.matrix(xn)))
  }
  auc_with <- as.numeric(pROC::auc(pROC::roc(y, p_with)))
  auc_no   <- as.numeric(pROC::auc(pROC::roc(y, p_no)))
  b_with <- mean((p_with - y)^2, na.rm = TRUE)
  b_no   <- mean((p_no - y)^2, na.rm = TRUE)
  return(c(auc_with = auc_with, auc_no = auc_no, b_with = b_with, b_no = b_no))
}

# Prepare test matrices for bootstrap (use numeric truth)
y_test_num <- ifelse(test_baked$target == "Yes", 1, 0)
# For RF bootstrap (caret models)
set.seed(2026)
B <- 500
boot_res_rf <- boot::boot(data = seq_len(nrow(x_test)), statistic = function(data, i){
  res <- compute_metrics(i, x_test, x_test_no_beh, y_test_num, rf_with, rf_no, model_type = "rf")
  c(res["auc_with"], res["auc_no"], res["b_with"], res["b_no"])
}, R = B)

# Compute delta distributions and 95% CIs
delta_auc_rf <- boot_res_rf$t[,1] - boot_res_rf$t[,2]
delta_brier_rf <- boot_res_rf$t[,3] - boot_res_rf$t[,4]
ci_auc_rf <- quantile(delta_auc_rf, probs = c(0.025, 0.5, 0.975), na.rm = TRUE)
ci_brier_rf <- quantile(delta_brier_rf, probs = c(0.025, 0.5, 0.975), na.rm = TRUE)

# XGB bootstrap (if direct models available)
if(exists("xgb_with") && exists("xgb_no")){
  set.seed(2026)
  boot_res_xgb <- boot::boot(data = seq_len(nrow(x_test)), statistic = function(data, i){
    res <- compute_metrics(i, x_test, x_test_no_beh, y_test_num, xgb_with, xgb_no, model_type = "xgb")
    c(res["auc_with"], res["auc_no"], res["b_with"], res["b_no"])
  }, R = B)
  delta_auc_xgb <- boot_res_xgb$t[,1] - boot_res_xgb$t[,2]
  delta_brier_xgb <- boot_res_xgb$t[,3] - boot_res_xgb$t[,4]
  ci_auc_xgb <- quantile(delta_auc_xgb, probs = c(0.025, 0.5, 0.975), na.rm = TRUE)
  ci_brier_xgb <- quantile(delta_brier_xgb, probs = c(0.025, 0.5, 0.975), na.rm = TRUE)
} else {
  ci_auc_xgb <- ci_brier_xgb <- NA
}

ci_tbl <- tibble(
  Model = c("RF delta (with - without)","XGB delta (with - without)"),
  AUC_delta_median = c(ci_auc_rf[2], ifelse(is.numeric(ci_auc_xgb), ci_auc_xgb[2], NA)),
  AUC_delta_95CI = c(paste0(round(ci_auc_rf[1],4)," to ",round(ci_auc_rf[3],4)), ifelse(is.numeric(ci_auc_xgb), paste0(round(ci_auc_xgb[1],4)," to ",round(ci_auc_xgb[3],4)), NA)),
  Brier_delta_median = c(ci_brier_rf[2], ifelse(is.numeric(ci_brier_xgb), ci_brier_xgb[2], NA)),
  Brier_delta_95CI = c(paste0(round(ci_brier_rf[1],6)," to ",round(ci_brier_rf[3],6)), ifelse(is.numeric(ci_brier_xgb), paste0(round(ci_brier_xgb[1],6)," to ",round(ci_brier_xgb[3],6)), NA))
)
knitr::kable(ci_tbl, caption = "Bootstrap CIs for ablation deltas (AUC and Brier): behavioral features effect")
```

---

# 14 Sanity checks and stability (permutation and repeated-split)

```{r sanity-stability}
# Permutation test (sanity check for leakage)
set.seed(2026)
# Align permutation with x_train rows
y_train_num <- ifelse(y_train == "Yes", 1L, 0L)
perm_y <- sample(y_train_num, size = length(y_train_num), replace = FALSE)
perm_dtrain <- xgboost::xgb.DMatrix(data = as.matrix(x_train), label = perm_y)
perm_cv <- tryCatch({
  xgboost::xgb.cv(params = if(exists("best_result")) best_result$params else list(objective="binary:logistic", eval_metric="auc", max_depth=3, eta=0.1, verbosity=0),
                  data = perm_dtrain, nrounds = if(exists("nrounds_final")) nrounds_final else 50,
                  nfold = 5, early_stopping_rounds = 10, verbose = FALSE, maximize = TRUE)
}, error = function(e) NULL)
if(!is.null(perm_cv)){
  auc_col <- grep("test_auc", names(perm_cv$evaluation_log), value = TRUE)[1]
  perm_auc <- tail(perm_cv$evaluation_log[[auc_col]], 1)
  message("Permutation CV AUC (sanity): ", perm_auc)
} else {
  message("Permutation CV could not be completed in this environment.")
}

# Repeated 5x70/30 splits stability for XGBoost (best-effort, small number of repeats)
set.seed(2026)
rep_n <- 5
stability <- tibble(rep = integer(), model = character(), auc = double())
for(r in seq_len(rep_n)){
  sp <- initial_split(model_data, prop = 0.7, strata = "target_attrit")
  tr <- training(sp); te <- testing(sp)
  tr_b <- bake(prep_rec, new_data = tr); te_b <- bake(prep_rec, new_data = te)
  xtr <- as.matrix(tr_b %>% select(any_of(names(x_train))))
  xte <- as.matrix(te_b %>% select(any_of(names(x_train))))
  # Build labels aligned with baked matrices
  ytr_num <- ifelse(tr_b$target == "Yes", 1L, 0L)
  yte_num <- ifelse(te_b$target == "Yes", 1L, 0L)
  dtr <- xgboost::xgb.DMatrix(data = xtr, label = ytr_num)
  mod <- tryCatch(xgboost::xgb.train(params = if(exists("best_result")) best_result$params else list(objective="binary:logistic", eval_metric="auc", max_depth=3, eta=0.1, verbosity=0),
                                    data = dtr, nrounds = if(exists("nrounds_final")) nrounds_final else 50, verbose = 0), error = function(e) NULL)
  if(!is.null(mod)){
    pred <- predict(mod, newdata = xgboost::xgb.DMatrix(data = xte))
    stability <- bind_rows(stability, tibble(rep = r, model = "xgb", auc = as.numeric(pROC::auc(pROC::roc(yte_num, pred)))))
  }
}
if(nrow(stability)>0) knitr::kable(stability, caption = "Repeated-split stability for XGBoost (AUC per repeat)") else message("Stability repeats did not produce results.")
```

---

# 15 A/B test design and ROI estimate (concise, operational)

```{r ab-test}
# Stakeholder assumption defaults (documented and adjustable)
assumed_absolute_lift <- 0.05            # absolute increase in retention among contacted customers (5 percentage points)
cost_per_outreach <- 5.00                # USD per customer contacted
expected_value_per_retained_customer <- 200.00  # USD expected lifetime value per retained customer

# Note printed to the report so readers know these are illustrative assumptions
assumption_note <- paste0(
  "Assumptions used for A/B test and ROI example: assumed_absolute_lift = ",
  assumed_absolute_lift, ", cost_per_outreach = $", cost_per_outreach,
  ", expected_value_per_retained_customer = $", expected_value_per_retained_customer,
  ". These are illustrative and should be updated with stakeholder inputs."
)
message(assumption_note)
cat("\n", assumption_note, "\n\n")

# Use calibrated probabilities (use_probs defined earlier)
use_probs <- if(exists("use_probs")) use_probs else if(exists("platt_pred")) platt_pred else raw_probs

# Example threshold and flag rate
threshold_example <- 0.10
flag_rate <- mean(use_probs >= threshold_example, na.rm = TRUE)

# Estimate baseline retention among flagged (approximate from test set)
p0 <- mean(test_baked$target_attrit[use_probs >= threshold_example] == 0, na.rm = TRUE)
p0 <- ifelse(is.na(p0), 0.5, p0)
p1 <- p0 + assumed_absolute_lift

# Power/sample size calculation for two-proportion test
alpha <- 0.05
power <- 0.8
es <- pwr::ES.h(p1, p0)
ss <- pwr::pwr.2p.test(h = es, sig.level = alpha, power = power)
n_per_arm <- ceiling(ss$n)

ab_test_plan <- tibble(
  Threshold = threshold_example,
  Flagged_fraction = round(flag_rate,4),
  Baseline_retention_among_flagged = round(p0,4),
  Expected_retention_with_treatment = round(p1,4),
  Sample_size_per_arm = n_per_arm,
  Total_sample_needed = n_per_arm * 2
)
knitr::kable(ab_test_plan, caption = "A/B test plan (example) for threshold 0.10; assumptions are illustrative")

# Simple ROI estimate (example)
num_contacted <- n_per_arm
expected_incremental_retained <- (p1 - p0) * num_contacted
expected_benefit <- expected_incremental_retained * expected_value_per_retained_customer
expected_cost <- num_contacted * cost_per_outreach
roi_estimate <- expected_benefit - expected_cost

roi_tbl <- tibble(
  Contacts = num_contacted,
  Expected_incremental_retained = round(expected_incremental_retained,2),
  Expected_benefit = round(expected_benefit,2),
  Expected_cost = round(expected_cost,2),
  Expected_net = round(roi_estimate,2)
)
knitr::kable(roi_tbl, caption = "Example ROI estimate for the A/B test (adjust costs and LTV to stakeholder inputs)")
```

```{r ab-test-sensitivity}
# Sensitivity table: vary lift, cost, and LTV and compute net per 1,000 contacts
lifts <- c(0.02, 0.05, 0.10)
costs <- c(2, 5, 10)
ltvs <- c(100, 200, 400)
sensitivity <- expand.grid(lift = lifts, cost = costs, ltv = ltvs) %>%
  as_tibble() %>%
  mutate(
    contacts = 1000,
    incremental_retained = lift * contacts,
    benefit = incremental_retained * ltv,
    cost_total = contacts * cost,
    net = benefit - cost_total
  ) %>%
  arrange(desc(net))
knitr::kable(sensitivity, digits = 2, caption = "Sensitivity of net benefit per 1,000 contacts across assumed lifts, costs, and LTVs")
```

> **Note (printed in the report):** The A/B test and ROI numbers above are illustrative. Replace `assumed_absolute_lift`, `cost_per_outreach`, and `expected_value_per_retained_customer` with stakeholder inputs to produce operational estimates.

---

# 16 Audit outputs, hyperparameters, and session info

```{r audit-summary}
# Show audit table (first 100 rows)
knitr::kable(head(audit_tbl, 100), caption = "Leakage audit (columns flagged and reasons)")

# Predictor audit summary (first 200 rows)
knitr::kable(head(predictor_audit, 200), caption = "Predictor audit (predictors used and suspicious flag)")

# Hyperparameter summary
hp_tbl <- tibble(
  Model = c("Random Forest (caret)","XGBoost (caret/direct)"),
  Tuned_Hyperparameters = c(
    paste0("mtry=", rf_model$bestTune$mtry, "; ntree=500; repeatedCV (5x5)"),
    paste0("caret best (if used): nrounds=", if(!is.null(xgb_caret$bestTune$nrounds)) xgb_caret$bestTune$nrounds else NA,
           ", max_depth=", if(!is.null(xgb_caret$bestTune$max_depth)) xgb_caret$bestTune$max_depth else NA,
           ", eta=", if(!is.null(xgb_caret$bestTune$eta)) xgb_caret$bestTune$eta else NA,
           "; direct xgb nrounds=", if(exists("nrounds_final")) nrounds_final else NA)
  ),
  Justification = c(
    "mtry controls feature subset per split; grid chosen to balance bias/variance; ntree=500 for stability.",
    "Grid focuses on depth and learning rate to control complexity; sampling fractions reduce overfitting; repeated CV used for stability; direct xgb trained with CV-selected nrounds."
  )
)
knitr::kable(hp_tbl, caption = "Hyperparameter tuning summary and justification")

sessionInfo()
```

---
