Below are the two updated R Markdown files with only the minimal repairs required to fix the knitting errors you reported. I did not change any of your report text or structure except for:

- In the EDA report I corrected the descriptive table selection to use the correct column name returned by `psych::describe` (`median` not `med`).
- In both reports I made kable captions safe for LaTeX by escaping underscores in captions that reference `test_baked`. This prevents the LaTeX "Missing $ inserted" error when captions contain underscores.
- I added a tiny helper `escape_caption()` used only for captions that include underscores. Nothing else was altered.

Copy each block into its respective `.Rmd` file and knit to PDF. The appendix code still attempts to load `artifacts/project_artifacts.RData` and will print clear diagnostics if artifacts are missing.

---

## 1) `EDA_Report_BruderG.Rmd`

```yaml
---
title: "Exploratory Data Analysis Report"
author: "Geoffrey Bruder"
date: "`r format(Sys.Date(), '%Y-%m-%d')`"
output:
  pdf_document:
    latex_engine: xelatex
    toc: true
    toc_depth: 2
geometry: margin=1in
fontsize: 12pt
---
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, fig.width = 6.5, fig.height = 4.5)
library(tidyverse)
library(knitr)
library(kableExtra)
library(ggplot2)
library(corrplot)
library(psych)
library(car)
library(pROC)
set.seed(2026)
# Attempt to load artifacts saved by the tuned project Rmd
artifacts_dir <- "artifacts"
artifacts_file <- file.path(artifacts_dir, "project_artifacts.RData")
if(file.exists(artifacts_file)){
  load(artifacts_file, envir = .GlobalEnv)
  message("Loaded project artifacts from: ", artifacts_file)
} else {
  # try rds fallbacks
  rds_try <- function(name){
    path <- file.path(artifacts_dir, paste0(name, ".rds"))
    if(file.exists(path)){
      assign(name, readRDS(path), envir = .GlobalEnv)
      message("Loaded ", name, " from ", path)
      TRUE
    } else FALSE
  }
  rds_try("test_baked")
  rds_try("xgb_pred_prob")
  rds_try("rf_pred_prob")
  rds_try("pred_base_prob")
  rds_try("pred_enh_prob")
  rds_try("cal_plot")
  rds_try("ci_tbl")
  message("project_artifacts.RData not found; attempted rds fallbacks. If artifacts are missing, run the tuned project Rmd to produce them.")
}

# helper to escape underscores in captions for LaTeX
escape_caption <- function(s) {
  gsub("_", "\\\\_", s)
}
```

Exploratory Data Analysis Report

This report documents the exploratory data analysis performed to support a predictive modeling pipeline for customer attrition. The analysis characterizes the BankChurners dataset used in the project, inspects distributions and relationships among candidate predictors, evaluates engineered behavioral‑delta features, and identifies data quality issues and preprocessing needs that informed the subsequent feature engineering and modeling work. The EDA emphasizes interpretable summaries and diagnostics that are directly relevant to stakeholder decisions about retention targeting and to reproducible model development.

# Background and research question

The project asks whether short‑term behavioral changes in customer activity—specifically quarter‑to‑quarter changes in transaction amount and transaction count and reductions in contact frequency—provide incremental predictive signal for attrition beyond static demographic and credit features. The working hypothesis is that customers exhibiting recent negative behavioral changes are more likely to attrite and that models incorporating these delta features will show improved discrimination and better calibrated probabilities compared with static‑feature baselines.

# Data sources and acquisition

The analysis uses the anonymized BankChurners dataset provided for the capstone. The dataset was delivered as a flat CSV with one row per customer and includes demographic fields, credit and product metrics, transaction aggregates, and precomputed quarter‑to‑quarter change measures. I preserved an anonymized PseudoID for export and removed direct identifiers during the leakage audit. No external scraping or API calls were required; all data used in EDA were those provided in the project dataset.

# Cleaning, merging, and aggregation prior to EDA

I standardized column names to snake_case, coerced numeric fields to numeric types, and consolidated categorical missingness into an explicit Unknown level. Records with missing or ambiguous target labels were removed from the modeling pool. I inspected each variable for implausible values and applied domain‑appropriate corrections where the provenance was clearly a formatting or import error (for example, trimming whitespace in categorical fields and coercing numeric strings). No external joins were necessary; the dataset was self‑contained. The targeted leakage audit removed any columns that appeared to be model‑derived scores or identifiers beyond the preserved PseudoID.

# EDA methods and rationale

Exploratory analysis combined univariate summaries, bivariate visualizations, and multivariate diagnostics. For continuous variables I computed median, mean, interquartile range, and skewness to capture central tendency and distributional shape; medians were emphasized where distributions were skewed. For categorical variables I computed absolute counts and proportions and inspected level cardinality to guide consolidation of rare levels. I produced at least five visual diagnostics that are most informative for modeling decisions: histograms and density plots for key continuous predictors, boxplots stratified by attrition for outcome‑related inspection, bar charts for categorical distributions, an ROC‑style inspection of raw score separability for engineered deltas, and a correlogram to assess pairwise linear associations and potential multicollinearity. I also computed a correlation matrix and visualized it as a heatmap to identify clusters of highly correlated predictors that might require dimensionality reduction or careful regularization in modeling. These choices prioritize interpretability for stakeholders and provide direct evidence to justify preprocessing decisions.

# Results: descriptive summaries and key visualizations

Descriptive summaries for the target and core predictors are provided in Appendix Table A1 and Appendix Table A2. The binary response, target_attrit, is derived from Attrition_Flag and is balanced enough for modeling after stratified splitting; exact class proportions are reported in Appendix Table A1. Continuous predictors such as Total_Trans_Amt and Avg_Utilization_Ratio show right skew; medians are therefore more representative than means for central tendency. The engineered delta features Amt_Chg_Q4Q1 and Ct_Chg_Q4Q1 have distributions concentrated near zero with a long left tail for large negative changes; the binary flags Large_Amt_Drop and Large_Ct_Drop capture the extreme left tail and occur in a minority of customers but are enriched among attrited customers. Categorical predictors such as Card_Category and Income_Category contain several rare levels; consolidation to an Unknown or Other level reduces sparsity without materially changing marginal distributions.

Figure 1 (histogram and density) shows the distribution of Total_Trans_Amt and highlights skew and a small number of high‑value outliers; Figure 2 (boxplots stratified by target) shows that median transaction amounts are lower for attrited customers, supporting the hypothesis that declining activity correlates with attrition. Figure 3 (bar chart) reports counts and proportions for Card_Category and Income_Category and documents the consolidation performed for rare levels. Figure 4 (scatter and marginal density) visualizes Amt_Chg_Q4Q1 against Ct_Chg_Q4Q1 and shows a positive association for customers with stable behavior and a cluster of customers with simultaneous drops in both amount and count. Figure 5 (correlogram) displays pairwise Pearson correlations among numeric predictors and identifies a small set of highly correlated features (for example, Total_Trans_Amt and Avg_Open_To_Buy) that informed later feature selection decisions. Brief interpretations accompany each figure in the appendix.

# Collinearity assessment

The correlation heatmap indicates several moderate to strong correlations among transaction‑level aggregates and between credit limit and utilization measures. No single pair of predictors exhibited near‑perfect correlation, but clusters of correlated variables suggest that regularized models or tree‑based ensembles will be preferable to unregularized linear models if the goal is stable predictive performance. The correlogram and variance inflation factor checks (reported in Appendix Table A3) guided the decision to remove zero‑variance predictors and to collapse redundant features where appropriate.

# Optional stratified summaries

I computed stratified summaries of Total_Trans_Amt and Contacts_Count_12_mon by attrition class. Attrited customers show lower median Total_Trans_Amt and fewer contacts in the prior 12 months, consistent with the behavioral‑delta hypothesis. These stratified checks are reported in Appendix Table A4 and were used to justify the inclusion of Recent_Contact_Reduction as an engineered binary feature.

# Discussion and next steps

The EDA supports the central hypothesis that recent negative behavioral changes are associated with higher attrition risk. The enhanced logistic model and ensemble experiments later confirmed that delta features improve discrimination and calibration, but the EDA was critical in identifying distributional issues, rare categorical levels, and correlated predictor clusters that required preprocessing. The immediate next steps are to implement the preprocessing recipe described below, to finalize the set of candidate predictors (targeting approximately 20–25 key features for initial modeling), and to prepare the stratified train/test split that will be used for repeated cross‑validation and holdout evaluation. The EDA also highlighted the need for a targeted leakage audit and for conservative handling of outliers and rare categories to avoid overfitting.

# Appendix: codebook and selected tables and figures

The appendix below attempts to render the codebook, descriptive tables, VIF diagnostics, correlation matrix, stratified summaries, and the five figures referenced above using artifacts exported from the tuned project R Markdown. If the artifacts are not present in `artifacts/project_artifacts.RData` or the `.rds` fallbacks, the code prints a clear diagnostic message and does not error.

```{r appendix-load-checks, echo=FALSE}
# Helper to check for required objects and print status
required_objs <- c("test_baked")
present <- required_objs %in% ls(envir = .GlobalEnv)
if(!all(present)){
  missing <- required_objs[!present]
  message("Warning: the following required objects are missing and some appendix outputs will be skipped: ", paste(missing, collapse = ", "))
} else {
  message("Appendix: required objects present. Proceeding to render tables and figures where possible.")
}
```

## Appendix Table A1: Codebook / Data Dictionary (generated from available data)

```{r appendix-codebook, echo=FALSE}
if(exists("test_baked", envir = .GlobalEnv)){
  df <- get("test_baked", envir = .GlobalEnv)
  codebook <- tibble(
    Variable = names(df),
    Type = sapply(df, function(x) class(x)[1]),
    Example = sapply(df, function(x) {
      if(is.numeric(x)) format(round(median(x, na.rm = TRUE),2)) else as.character(head(unique(x),1))
    })
  )
  cap <- escape_caption("Appendix Table A1. Codebook and data dictionary (derived from test_baked).")
  knitr::kable(codebook, caption = cap, format = "latex", escape = TRUE) %>%
    kable_styling(latex_options = c("hold_position"))
} else {
  cat("Appendix Table A1 could not be rendered: test_baked is not available in the environment. Save artifacts from the tuned project Rmd and place them in artifacts/project_artifacts.RData.")
}
```

## Appendix Table A2: Univariate descriptive statistics for selected continuous predictors

```{r appendix-descriptives, echo=FALSE}
if(exists("test_baked", envir = .GlobalEnv)){
  df <- get("test_baked", envir = .GlobalEnv)
  cont_vars <- names(df)[sapply(df, is.numeric)]
  # select a subset for display (top 20 or all if fewer)
  cont_vars_sel <- head(cont_vars, 20)
  desc_tbl <- psych::describe(df[cont_vars_sel]) %>%
    as.data.frame() %>%
    rownames_to_column("Variable") %>%
    select(Variable, n, mean, sd, median = median, min, max, skew)
  knitr::kable(desc_tbl, caption = "Appendix Table A2. Descriptive statistics for selected continuous predictors.", format = "latex") %>%
    kable_styling(latex_options = c("hold_position"))
} else {
  cat("Appendix Table A2 could not be rendered: test_baked is not available.")
}
```

## Appendix Table A3: Correlation matrix and VIF diagnostics

```{r appendix-corr-vif, echo=FALSE, fig.width=6.5, fig.height=6}
if(exists("test_baked", envir = .GlobalEnv)){
  df <- get("test_baked", envir = .GlobalEnv)
  num_df <- df %>% select(where(is.numeric))
  if(ncol(num_df) >= 2){
    cor_mat <- cor(num_df, use = "pairwise.complete.obs")
    # print small correlation table
    cor_tbl <- as.data.frame(round(cor_mat, 2))
    knitr::kable(cor_tbl[1:min(10,nrow(cor_tbl)), 1:min(10,ncol(cor_tbl))], caption = "Appendix Table A3 (partial). Correlation matrix (first 10 numeric variables).", format = "latex") %>%
      kable_styling(latex_options = c("hold_position"))
    # plot correlogram
    corrplot::corrplot(cor_mat, method = "color", tl.cex = 0.6, number.cex = 0.6, main = "Correlogram (numeric predictors)")
    # VIF: fit a simple linear model on a subset to compute VIFs (use target as numeric if present)
    if("target_attrit" %in% names(df)){
      # create a small model for VIF diagnostics using top numeric predictors
      lm_vars <- names(num_df)[1:min(10, ncol(num_df))]
      lm_df <- num_df[lm_vars]
      # add a small constant if perfect collinearity
      lm_fit <- tryCatch(lm(lm_df[[1]] ~ ., data = lm_df[-1]), error = function(e) NULL)
      if(!is.null(lm_fit)){
        vif_vals <- car::vif(lm_fit)
        vif_tbl <- tibble(Variable = names(vif_vals), VIF = as.numeric(vif_vals))
        knitr::kable(vif_tbl, caption = "Appendix Table A3b. VIF diagnostics for a representative numeric subset.", format = "latex") %>%
          kable_styling(latex_options = c("hold_position"))
      } else {
        cat("VIF diagnostics skipped: unable to fit representative linear model for VIF calculation.")
      }
    }
  } else {
    cat("Not enough numeric variables to compute correlation matrix.")
  }
} else {
  cat("Appendix Table A3 could not be rendered: test_baked is not available.")
}
```

## Appendix Table A4: Stratified summaries by attrition class

```{r appendix-stratified, echo=FALSE}
if(exists("test_baked", envir = .GlobalEnv)){
  df <- get("test_baked", envir = .GlobalEnv)
  if("target_attrit" %in% names(df)){
    # choose a few key continuous vars to stratify
    vars <- c("Total_Trans_Amt","Total_Trans_Ct","Contacts_Count_12_mon","Avg_Utilization_Ratio")
    vars <- vars[vars %in% names(df)]
    strat_tbl <- df %>%
      mutate(target_num = ifelse(as.character(target_attrit) %in% c("Yes","1","TRUE"), 1, 0)) %>%
      group_by(target_attrit) %>%
      summarize(across(all_of(vars), list(median = ~median(.x, na.rm = TRUE), mean = ~mean(.x, na.rm = TRUE)), .names = "{col}_{fn}"), .groups = "drop")
    knitr::kable(strat_tbl, caption = "Appendix Table A4. Stratified summaries (median and mean) for selected predictors by attrition class.", format = "latex") %>%
      kable_styling(latex_options = c("hold_position"))
  } else {
    cat("Appendix Table A4 could not be rendered: target_attrit not found in test_baked.")
  }
} else {
  cat("Appendix Table A4 could not be rendered: test_baked is not available.")
}
```

## Appendix Figures 1–5 (histogram, boxplot, bar chart, scatter, correlogram)

```{r appendix-figures, echo=FALSE, fig.width=6.5, fig.height=4.5}
if(exists("test_baked", envir = .GlobalEnv)){
  df <- get("test_baked", envir = .GlobalEnv)
  # Figure 1: Total_Trans_Amt histogram + density
  if("Total_Trans_Amt" %in% names(df)){
    print(
      ggplot(df, aes(x = Total_Trans_Amt)) +
      geom_histogram(aes(y = ..density..), bins = 60, fill = "grey80", color = "black") +
      geom_density(color = "blue", size = 0.8) +
      labs(title = "Figure 1. Distribution of Total_Trans_Amt", x = "Total_Trans_Amt", y = "Density") +
      theme_minimal()
    )
  } else {
    plot.new(); text(0.5,0.5,"Figure 1 not available: Total_Trans_Amt missing.")
  }
  # Figure 2: boxplot Total_Trans_Amt by target
  if(all(c("Total_Trans_Amt","target_attrit") %in% names(df))){
    print(
      ggplot(df, aes(x = as.factor(target_attrit), y = Total_Trans_Amt)) +
      geom_boxplot() +
      labs(title = "Figure 2. Total_Trans_Amt by Attrition", x = "Attrition", y = "Total_Trans_Amt") +
      theme_minimal()
    )
  } else {
    plot.new(); text(0.5,0.5,"Figure 2 not available: required variables missing.")
  }
  # Figure 3: bar chart for Card_Category
  if("Card_Category" %in% names(df)){
    print(
      ggplot(df, aes(x = fct_lump(Card_Category, n = 6))) +
      geom_bar(fill = "steelblue") +
      labs(title = "Figure 3. Card_Category counts (lumped)", x = "Card_Category (lumped)", y = "Count") +
      theme_minimal()
    )
  } else {
    plot.new(); text(0.5,0.5,"Figure 3 not available: Card_Category missing.")
  }
  # Figure 4: scatter Amt_Chg_Q4Q1 vs Ct_Chg_Q4Q1
  if(all(c("Amt_Chg_Q4Q1","Ct_Chg_Q4Q1") %in% names(df))){
    print(
      ggplot(df, aes(x = Amt_Chg_Q4Q1, y = Ct_Chg_Q4Q1)) +
      geom_point(alpha = 0.4) +
      geom_smooth(method = "loess", se = FALSE, color = "red") +
      labs(title = "Figure 4. Amt_Chg_Q4Q1 vs Ct_Chg_Q4Q1", x = "Amt_Chg_Q4Q1", y = "Ct_Chg_Q4Q1") +
      theme_minimal()
    )
  } else {
    plot.new(); text(0.5,0.5,"Figure 4 not available: delta features missing.")
  }
  # Figure 5: correlogram already printed above via corrplot; if not, print a small placeholder
  if(exists("cor_mat")){
    # corrplot already drawn above; nothing to do
    NULL
  } else {
    plot.new(); text(0.5,0.5,"Figure 5 (correlogram) not available: numeric correlation matrix missing.")
  }
} else {
  plot.new(); text(0.5,0.5,"Appendix figures could not be rendered: test_baked not available.")
}
```

# Repository and reproducibility

All code used to generate the EDA tables and figures is available in the project R Markdown files in the repository. If you need the artifacts used to render these appendix outputs, run the tuned project R Markdown to produce `artifacts/project_artifacts.RData` and place it in the `artifacts/` folder adjacent to this R Markdown.
---
