---
title: "Baseline Models Report — County Childhood Vaccination Uptake"
author: "Geoffrey Bruder"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    toc: true
    toc_depth: 3
    df_print: paged
  pdf_document:
    latex_engine: xelatex
    toc: true
    toc_depth: 3
    number_sections: true
params: null
---
```

```{r setup_m05, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, fig.width = 7, fig.height = 4.5)
library(tidyverse)
library(jsonlite)
library(kableExtra)
library(scales)
library(glue)
library(here)
proj_root <- here::here()
outputs_dir <- file.path(proj_root, "outputs")
models_dir <- file.path(proj_root, "models")
dir.create(outputs_dir, showWarnings = FALSE, recursive = TRUE)
dir.create(models_dir, showWarnings = FALSE, recursive = TRUE)

read_models_summary <- function() {
  p <- file.path(outputs_dir, "models_summary.json")
  if (file.exists(p)) jsonlite::fromJSON(p) else NULL
}
ms <- read_models_summary()
```

# Executive summary

This Baseline Models report summarizes initial model selection, nested CV results, residual diagnostics, formal assumptions tests, and interpretability outputs (coefficients and SHAP). The goal is to present baseline performance and evidence of overfitting control.

\newpage

## Model comparison — nested CV results

```{r model-summary-table, echo=TRUE}
if (!is.null(ms)) {
  models <- ms[names(ms) != "run_info"]
  df_models <- map_dfr(names(models), function(nm) {
    v <- models[[nm]]
    tibble(
      model = nm,
      cv_rmse_mean = v$cv_rmse_mean %||% NA,
      cv_rmse_std = v$cv_rmse_std %||% NA,
      cv_mae_mean = v$cv_mae_mean %||% NA,
      cv_r2_mean = v$cv_r2_mean %||% NA
    )
  })
  df_models %>%
    arrange(cv_rmse_mean) %>%
    mutate(across(where(is.numeric), ~ round(.x, 3))) %>%
    kable(caption = "Nested CV model comparison (lower RMSE better)") %>%
    kable_styling(full_width = FALSE)
} else {
  cat("models_summary.json not found in outputs/ — run nested CV cell first.\n")
}
```

```{r model-rmse-plot, fig.height=4.5}
if (!is.null(ms)) {
  df_plot <- df_models %>% arrange(cv_rmse_mean)
  ggplot(df_plot, aes(x = reorder(model, cv_rmse_mean), y = cv_rmse_mean)) +
    geom_col(fill = "#2c7fb8", alpha = 0.9) +
    geom_errorbar(aes(ymin = cv_rmse_mean - cv_rmse_std, ymax = cv_rmse_mean + cv_rmse_std), width = 0.2) +
    coord_flip() +
    labs(x = "", y = "CV RMSE (mean ± std)", caption = "Nested CV results; lower RMSE indicates better predictive performance") +
    theme_minimal(base_size = 11)
} else {
  cat("No model summary available to plot.\n")
}
```

\newpage

## Best model selection and artifact check

```{r best-model, echo=TRUE}
if (!is.null(ms)) {
  df_models2 <- df_models %>% arrange(cv_rmse_mean)
  best <- df_models2$model[1]
  cat("Best model by nested CV RMSE:", best, "\n\n")
  best_info <- ms[[best]]
  print(best_info$best_params_full %||% "No best_params_full recorded")
  manifest_path <- file.path(outputs_dir, paste0(best, "_artifact_manifest.json"))
  if (file.exists(manifest_path)) {
    print(jsonlite::fromJSON(manifest_path))
  } else {
    cat("No artifact manifest found for", best, "\n")
  }
} else {
  cat("No models_summary.json found.\n")
}
```

\newpage

## Training vs CV performance (baseline model)

```{r train-vs-cv, echo=TRUE}
metrics_path <- file.path(outputs_dir, "training_vs_cv_metrics.csv")
if (file.exists(metrics_path)) {
  readr::read_csv(metrics_path, show_col_types = FALSE) %>% mutate(across(where(is.numeric), ~ round(.x,3))) %>% kable(caption="Training vs CV performance (baseline models)") %>% kable_styling(full_width = FALSE)
} else {
  cat("training_vs_cv_metrics.csv not found. The notebook can produce this by refitting models on training folds and saving metrics.\n")
}
```

\newpage

## Residual diagnostics for the selected baseline

```{r residuals, fig.height=4.5}
if (!is.null(ms)) {
  best <- df_models %>% arrange(cv_rmse_mean) %>% slice(1) %>% pull(model)
  preds_path <- file.path(outputs_dir, paste0(best, "_preds_full.csv"))
  if (file.exists(preds_path)) {
    preds <- readr::read_csv(preds_path, show_col_types = FALSE)
    preds <- preds %>% mutate(resid = y_true - y_pred)
    p1 <- ggplot(preds, aes(x = y_pred, y = resid)) +
      geom_point(alpha = 0.5) +
      geom_hline(yintercept = 0, color = "red", linetype = "dashed") +
      labs(x = "Predicted", y = "Residual", caption = glue("Residuals vs predicted for {best}")) +
      theme_minimal(base_size = 11)
    p2 <- ggplot(preds, aes(x = resid)) + geom_histogram(bins = 40, fill = "#2b8cbe", color = "white") + labs(x = "Residual", y = "Count", caption = "Residual histogram") + theme_minimal(base_size = 11)
    print(p1); print(p2)
  } else {
    cat("Predictions file not found for", best, "\n")
  }
} else {
  cat("No models_summary.json found.\n")
}
```

\newpage

## Formal assumptions tests

```{r assumptions, echo=TRUE}
if (!is.null(ms)) {
  best <- df_models %>% arrange(cv_rmse_mean) %>% slice(1) %>% pull(model)
  preds_path <- file.path(outputs_dir, paste0(best, "_preds_full.csv"))
  if (file.exists(preds_path)) {
    preds <- readr::read_csv(preds_path, show_col_types = FALSE) %>% mutate(resid = y_true - y_pred)
    if (nrow(preds) <= 5000) {
      sh <- shapiro.test(preds$resid)
      cat("Shapiro-Wilk p-value:", round(sh$p.value, 4), "\n")
      cat("Interpretation: p < 0.05 suggests residuals deviate from normality; with large N, small deviations can be significant.\n")
    } else {
      cat("Shapiro-Wilk not run (n > 5000). Use histogram and skew/kurtosis diagnostics instead.\n")
      if (requireNamespace("e1071", quietly = TRUE)) {
        cat("Residual skew:", round(e1071::skewness(preds$resid, na.rm = TRUE), 3), "kurtosis:", round(e1071::kurtosis(preds$resid, na.rm = TRUE), 3), "\n")
      } else {
        cat("Package 'e1071' not installed; cannot compute skew/kurtosis.\n")
      }
    }
    if (requireNamespace("lmtest", quietly = TRUE)) {
      lm_fit <- lm(y_true ~ y_pred, data = preds)
      bp <- lmtest::bptest(lm_fit)
      cat("Breusch-Pagan p-value:", round(bp$p.value, 4), "\n")
      cat("Interpretation: p < 0.05 indicates heteroskedasticity (non-constant variance).\n")
    } else {
      cat("Package 'lmtest' not installed; cannot run Breusch-Pagan test.\n")
    }
    cat("Durbin-Watson test: only applicable if data are time-ordered. If no time index exists, this test is not applicable.\n")
  } else {
    cat("Predictions file not found for", best, "; cannot run formal residual tests.\n")
  }
} else {
  cat("No models_summary.json found.\n")
}
```

\newpage

## ElasticNet coefficients and interpretation

```{r elasticnet-coefs, echo=TRUE}
# Use file.path and coerce coefficient column to numeric before abs()
coef_path <- file.path(outputs_dir, "elasticnet_coefficients.csv")
if (file.exists(coef_path)) {
  coefs <- readr::read_csv(coef_path, show_col_types = FALSE)
  # Ensure coefficient column exists and is numeric
  if (!"coef" %in% names(coefs)) {
    # try common alternative names
    possible <- intersect(c("coefficient", "estimate", "beta"), names(coefs))
    if (length(possible) > 0) {
      coefs <- coefs %>% rename(coef = !!sym(possible[1]))
    } else {
      stop("No coefficient column found in elasticnet_coefficients.csv")
    }
  }
  coefs <- coefs %>% mutate(coef_numeric = suppressWarnings(as.numeric(coef)))
  # If conversion produced NAs, warn and attempt to parse commas/strings
  if (any(is.na(coefs$coef_numeric) & !is.na(coefs$coef))) {
    coefs <- coefs %>% mutate(coef_numeric = as.numeric(gsub(",", "", as.character(coef))))
  }
  coefs %>%
    arrange(desc(abs(coef_numeric))) %>%
    mutate(coef = round(coef_numeric, 4)) %>%
    slice_head(n = 10) %>%
    select(-coef_numeric) %>%
    kable(caption = "Top 10 ElasticNet coefficients (by absolute value)") %>%
    kable_styling(full_width = FALSE)
  cat("Interpretation: positive coefficients increase predicted coverage; negative coefficients decrease predicted coverage, holding other features constant.\n")
} else {
  cat("ElasticNet coefficients file not found. The notebook can export coefficients to outputs/elasticnet_coefficients.csv.\n")
}
```

\newpage

## SHAP directionality / partial dependence for top features

```{r shap-direction, echo=TRUE}
sv_path <- file.path(outputs_dir, "shap_values_random_forest_sample.csv")
if (file.exists(sv_path)) {
  sv <- readr::read_csv(sv_path, show_col_types = FALSE)
  feat_cols <- names(sv)[grepl('feat_', names(sv))]
  if (length(feat_cols) > 0) {
    ma <- sv %>% select(all_of(feat_cols)) %>% summarise(across(everything(), ~ mean(abs(.x), na.rm = TRUE))) %>% pivot_longer(everything(), names_to = 'feature', values_to = 'mean_abs_shap') %>% arrange(desc(mean_abs_shap)) %>% slice_head(n = 3)
    ma %>% mutate(mean_abs_shap = round(mean_abs_shap, 4)) %>% kable(caption = "Top 3 features by mean absolute SHAP (sample)") %>% kable_styling(full_width = FALSE)
    cat("For each top feature, SHAP directionality plots are available in outputs/ if generated by the notebook. Use SHAP dependence plots to interpret directionality.\n")
  } else {
    cat("No feature columns matching 'feat_' found in SHAP CSV.\n")
  }
} else {
  cat("Random forest SHAP CSV not found; skipping SHAP top features table.\n")
}
```

\newpage

## Overfitting control summary

- Nested cross‑validation (outer 5‑fold, inner 3‑fold) was used to estimate generalization performance and tune hyperparameters.  
- Regularization: ElasticNet uses L1/L2 penalties; tuned `alpha` and `l1_ratio`.  
- Tree constraints: `max_depth` and `max_features` were included in search spaces to limit overfitting.  
- MLP: early stopping / patience implemented in the PyTorch wrapper to avoid overfitting.  
- Next step: temporal holdout evaluation if time-indexed data is available.

\newpage

## Appendix — reproducibility and artifacts

- Model artifacts: `models/{model}_final.joblib`  
- Predictions: `outputs/{model}_preds_full.csv`  
- Coefficients: `outputs/elasticnet_coefficients.csv`  
- Training vs CV metrics: `outputs/training_vs_cv_metrics.csv`  
- SHAP CSVs: `outputs/shap_values_*.csv`  
- For PDF knitting: ensure `lmtest`, `e1071`, `kableExtra` installed.
```
