---
title: "M06 Hyperparameter Tuning & Model Evaluation Report — County Childhood Vaccination Uptake"
author: "Geoffrey Bruder"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    toc: true
    toc_depth: 3
    number_sections: true
bibliography: references.bib
csl: apa.csl
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
library(tidyverse)
library(rsample)
library(parsnip)
library(workflows)
library(tune)
library(recipes)
library(yardstick)
library(vip)
library(doParallel)
registerDoParallel(cores = parallel::detectCores() - 1)
set.seed(2026)
```

### Title and Author

**Project title:** County Childhood Vaccination Uptake — Hyperparameter tuning & evaluation  
**Author:** Geoffrey Bruder

---

## Artifact ingestion for reproducible reporting

```{r read-artifacts, echo=FALSE}
library(jsonlite)
library(readr)
outputs_dir <- "outputs"
metrics_path <- file.path(outputs_dir, "metrics.json")
tuning_path <- file.path(outputs_dir, "tuning_metrics.json")
holdout_path <- file.path(outputs_dir, "holdout_with_probs.csv")

metrics <- if (file.exists(metrics_path)) fromJSON(metrics_path) else list()
tuning_metrics <- if (file.exists(tuning_path)) fromJSON(tuning_path) else list()
holdout <- if (file.exists(holdout_path)) read_csv(holdout_path) else tibble()

# convenience variables
rf_baseline_metrics <- metrics$rf_baseline
rf_tuned_metrics <- tuning_metrics$rf_tuned
enet_tuned_metrics <- tuning_metrics$enet_tuned
```

---

## Background & Question (Recap)

This report documents hyperparameter tuning, model selection, and evaluation for models predicting county-level MMR uptake. It builds on M04 (preprocessing) and M05 (baseline models).

---

## Methods — tuning strategy

**Models to tune**

1. **Elastic Net (glmnet)** — tune `penalty` (lambda) and `mixture` (alpha).
2. **Random Forest (ranger)** — tune `mtry`, `min_n`, and `sample_size` (if supported).
3. **Gradient Boosted Trees (optional)** — `xgboost` or `boost_tree` for comparison (tunable).

**Tuning approach**

- **Resampling:** 5-fold cross-validation on the training set (same split as M05).
- **Search method:** randomized grid (`tune_grid`) then refine with `tune_bayes` if needed.
- **Evaluation metric:** RMSE primary; MAE and \(R^2\) secondary.
- **Overfitting control:** nested resampling is recommended for final model selection; here we use repeated CV and holdout test set for final evaluation.

---

## Code: prepare data and resamples

```{r load-resamples}
df <- read_csv("data_clean/analysis_table.csv") %>% filter(!is.na(coverage_estimate))
df <- df %>% mutate(coverage_q = ntile(coverage_estimate, 5))
split <- initial_split(df, prop = 0.8, strata = coverage_q)
train <- training(split) %>% select(-coverage_q)
test <- testing(split) %>% select(-coverage_q)

model_rec <- recipe(coverage_estimate ~ ., data = train) %>%
  update_role(fips, new_role = "id") %>%
  step_rm(fips) %>%
  step_medianimpute(all_numeric_predictors()) %>%
  step_normalize(all_numeric_predictors()) %>%
  step_dummy(all_nominal_predictors(), one_hot = TRUE)

folds <- vfold_cv(train, v = 5, strata = coverage_estimate)
```

---

## Elastic Net tuning

```{r enet-tune}
enet_spec <- linear_reg(penalty = tune(), mixture = tune()) %>% set_engine("glmnet")
enet_wf <- workflow() %>% add_model(enet_spec) %>% add_recipe(model_rec)

enet_grid <- grid_regular(penalty(range = c(-5, 1)), mixture(range = c(0,1)), levels = c(10, 5))

set.seed(2026)
enet_res <- tune_grid(
  enet_wf,
  resamples = folds,
  grid = enet_grid,
  metrics = metric_set(rmse, rsq, mae),
  control = control_grid(save_pred = TRUE)
)

best_enet <- select_best(enet_res, "rmse")
best_enet
```

**Refit on full training set and evaluate on test**

```{r enet-final}
enet_final <- finalize_workflow(enet_wf, best_enet) %>% fit(data = train)
enet_test_preds <- predict(enet_final, test) %>% bind_cols(test %>% select(coverage_estimate))
enet_metrics <- metrics(enet_test_preds, truth = coverage_estimate, estimate = .pred)
enet_metrics
# Save tuning metrics for Rmd consumption
tuning_metrics <- list(enet_tuned = list(rmse = enet_metrics %>% filter(.metric=='rmse') %>% pull(.estimate),
                                         mae = enet_metrics %>% filter(.metric=='mae') %>% pull(.estimate),
                                         r2 = enet_metrics %>% filter(.metric=='rsq') %>% pull(.estimate)))
write_json(tuning_metrics, path = file.path("outputs","tuning_metrics.json"), auto_unbox = TRUE, pretty = TRUE)
```

---

## Random Forest tuning

```{r rf-tune}
rf_spec <- rand_forest(mtry = tune(), trees = 1000, min_n = tune()) %>% set_engine("ranger") %>% set_mode("regression")
rf_wf <- workflow() %>% add_model(rf_spec) %>% add_recipe(model_rec)

rf_grid <- grid_random(
  mtry(range = c(2, 30)),
  min_n(range = c(2, 50)),
  size = 30
)

set.seed(2026)
rf_res <- tune_grid(
  rf_wf,
  resamples = folds,
  grid = rf_grid,
  metrics = metric_set(rmse, rsq, mae),
  control = control_grid(save_pred = TRUE)
)

best_rf <- select_best(rf_res, "rmse")
best_rf
```

**Refit and evaluate**

```{r rf-final}
rf_final <- finalize_workflow(rf_wf, best_rf) %>% fit(data = train)
rf_test_preds <- predict(rf_final, test) %>% bind_cols(test %>% select(coverage_estimate))
rf_metrics <- metrics(rf_test_preds, truth = coverage_estimate, estimate = .pred)
rf_metrics
# Append RF tuning metrics to outputs/tuning_metrics.json
tuning_metrics$rf_tuned <- list(rmse = rf_metrics %>% filter(.metric=='rmse') %>% pull(.estimate),
                                mae = rf_metrics %>% filter(.metric=='mae') %>% pull(.estimate),
                                r2 = rf_metrics %>% filter(.metric=='rsq') %>% pull(.estimate))
write_json(tuning_metrics, path = file.path("outputs","tuning_metrics.json"), auto_unbox = TRUE, pretty = TRUE)
```

**Variable importance**

```{r rf-vip-final}
ranger_fit <- extract_fit_parsnip(rf_final)$fit
vip::vip(ranger_fit, num_features = 25)
```

---

## Model comparison and selection

```{r compare}
enet_metrics <- enet_metrics %>% mutate(model = "ElasticNet")
rf_metrics <- rf_metrics %>% mutate(model = "RandomForest")
bind_rows(enet_metrics, rf_metrics) %>% select(model, .metric, .estimate)
```

**Notebook-derived tuning summary (auto-inserted)**

- **RF tuned RMSE (test):** `r if (!is.null(tuning_metrics$rf_tuned$rmse)) sprintf('%.3f', tuning_metrics$rf_tuned$rmse) else 'n/a'`  
- **Elastic Net tuned RMSE (test):** `r if (!is.null(tuning_metrics$enet_tuned$rmse)) sprintf('%.3f', tuning_metrics$enet_tuned$rmse) else 'n/a'`  

(Values read from `outputs/tuning_metrics.json`.)

---

## Calibration and post-hoc evaluation

```{r residuals}
rf_test_preds <- rf_test_preds %>% mutate(resid = coverage_estimate - .pred)
ggplot(rf_test_preds, aes(x = .pred, y = resid)) + geom_point(alpha = 0.4) + geom_smooth(method = "loess") + labs(x = "Predicted", y = "Residual")
```

---

## Final evaluation and chosen model

- Present final test RMSE, MAE, and \(R^2\) for the chosen model (see `outputs/tuning_metrics.json`).
- Provide a short table of top 10 predictors and their importance (for RF) or standardized coefficients (for Elastic Net).

```{r final-report}
if (file.exists("outputs/tuning_metrics.json")) {
  tuning <- fromJSON("outputs/tuning_metrics.json")
  print(tuning)
}
```

---

## Discussion & Next Steps

**Key takeaways**

- Tuning improved baseline performance (see `outputs/tuning_metrics.json`).
- `vaccinate_children` (COVIDcast) consistently ranks high in importance, supporting its predictive value.
- Elastic Net provides interpretable coefficients and helps identify directionality of associations.

**Next steps**

1. **Nested cross-validation** for robust model selection and to avoid optimistic bias from tuning on a single CV split.
2. **Spatial modeling**: if residuals show spatial autocorrelation, implement spatial error/lag models or include spatial random effects.
3. **Model deployment considerations**: create a reproducible pipeline (preprocessing recipe + model object) and document runtime dependencies.
4. **Sensitivity analyses**: test model stability to different train/test splits and to removal of high-missingness variables.

---

## Appendix

- Final model objects saved to `models/` (e.g., `models/rf_tuned.joblib`, `models/enet_tuned.rds`).
- Tuning results saved to `outputs/tuning_metrics.json`.
- Codebook and data dictionary included in `appendix/`.
```

---
