---
title: "Hyperparameter Tuning & Model Evaluation Report — County Childhood Vaccination Uptake"
author: "Geoffrey Bruder"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    toc: true
    toc_depth: 3
    df_print: paged
  pdf_document:
    latex_engine: xelatex
    toc: true
    toc_depth: 3
    number_sections: true
params: null
---

```{r setup_m06, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, fig.width=7, fig.height=4.5)
library(tidyverse)
library(jsonlite)
library(kableExtra)
library(scales)
library(glue)
library(fs)
library(here)
proj_root <- here::here()
outputs_dir <- path(proj_root, "outputs")
models_dir <- path(proj_root, "models")
dir_create(outputs_dir)
dir_create(models_dir)
ms_path <- path(outputs_dir, "models_summary.json")
ms <- if (file_exists(ms_path)) fromJSON(ms_path) else NULL
```

# Executive summary

This report documents hyperparameter tuning strategy, final tuned hyperparameters, conformal intervals and calibration diagnostics, and recommendations for further tuning and deployment.

\newpage

## Tuning methodology (recap)

- Nested cross‑validation: outer 5‑fold CV for performance estimation; inner 3‑fold CV for hyperparameter search (RandomizedSearchCV).  
- Randomized search balances exploration and compute; parameter validation prevents invalid parameter choices.  
- Final refit: best hyperparameters re-searched on full cleaned dataset and final models saved.

\newpage

## Tuned hyperparameters (parsed table)

```{r hyperparam-table, echo=TRUE}
if (!is.null(ms)) {
  rows <- map_dfr(names(ms)[names(ms) != "run_info"], function(nm) {
    v <- ms[[nm]]
    bp <- v$best_params_full %||% list()
    tibble(
      model = nm,
      n_estimators = bp$n_estimators %||% NA,
      max_depth = bp$max_depth %||% NA,
      learning_rate = bp$learning_rate %||% NA,
      num_leaves = bp$num_leaves %||% NA,
      alpha = bp$alpha %||% NA,
      l1_ratio = bp$l1_ratio %||% NA
    )
  })
  rows %>% mutate(across(where(is.numeric), ~ round(.x, 4))) %>% kable(caption="Parsed best hyperparameters per model (final refit)") %>% kable_styling(full_width = FALSE)
} else {
  cat("models_summary.json not found; run nested CV cell first.\n")
}
```

**Figure 1. Parsed best hyperparameters per model (table).**  
*Description.* Table showing the most important tuned hyperparameters for each model after final refit (e.g., `n_estimators`, `max_depth`, `learning_rate`, `alpha`, `l1_ratio`). *Interpretation.* The table shows ElasticNet tuned to a high `alpha` and high `l1_ratio` (strong regularization with L1 emphasis), Random Forest using many trees and moderate depth, and GBM using a small learning rate with many estimators — consistent with standard bias/variance tradeoffs.

\newpage

## Hyperparameter search space justification

```{r hyperparam-just, echo=TRUE}
cat("Justifications (concise):\n\n")
cat("- Random Forest: n_estimators 100–600 to balance variance reduction vs compute; max_depth options to control tree complexity and avoid overfitting; max_features choices to control feature sampling.\n\n")
cat("- Gradient Boosting (sklearn GBM / LightGBM): learning_rate 0.01–0.2 to trade off bias/variance; n_estimators wide range to allow many small steps; max_depth and num_leaves to control complexity.\n\n")
cat("- ElasticNet: alpha and l1_ratio tuned to balance regularization strength and L1/L2 mix for sparsity and stability.\n\n")
cat("- MLP (PyTorch): hidden_sizes and learning rate tuned; early stopping/patience used to avoid overfitting; weight decay for regularization.\n")
```

\newpage

## Calibration plot and reliability diagram

```{r calibration-plot, echo=TRUE, fig.height=4.5}
cal_files <- if (dir_exists(outputs_dir)) dir_ls(outputs_dir, regexp = "calibration_.*\\.csv", recurse = FALSE) else character(0)
if (length(cal_files) > 0) {
  for (cf in cal_files) {
    dfc <- read_csv(cf, show_col_types = FALSE)
    model_name <- str_remove(path_file(cf), "calibration_|\\.csv")
    p <- ggplot(dfc, aes(x = predicted_quantile, y = observed_quantile)) +
      geom_line() + geom_abline(slope=1, intercept=0, linetype="dashed", color="gray") +
      labs(x="Predicted quantile", y="Observed quantile", caption = glue("Calibration plot for {model_name}")) +
      theme_minimal(base_size=11)
    print(p)
  }
} else {
  cat("No calibration CSVs found in outputs/. The notebook can compute predicted vs observed quantiles and save to outputs/calibration_{model}.csv.\n")
}
```

**Figure 2. Calibration plots (predicted vs observed quantiles).**  
*Description.* Line plots comparing predicted quantiles (median predicted values by bin) to observed quantiles (median observed values) for each model. A 45° dashed line indicates perfect calibration. *Interpretation.* If a model's line lies close to the diagonal, predicted quantiles match observed quantiles and the model is well calibrated. In the current run no calibration CSVs were found, so calibration could not be plotted; if produced, these plots would show whether models systematically over‑ or under‑predict across the distribution.

\newpage

## Conformal coverage diagnostic

```{r conformal-coverage, echo=TRUE}
if (!is.null(ms)) {
  cov_rows <- map_dfr(names(ms)[names(ms) != "run_info"], function(nm) {
    v <- ms[[nm]]
    q90 <- v$conformal_q_90 %||% NA
    preds_path <- path(outputs_dir, paste0(nm, "_preds_full.csv"))
    if (file_exists(preds_path) && !is.na(q90)) {
      preds <- read_csv(preds_path, show_col_types = FALSE)
      covered <- mean((preds$y_true >= (preds$y_pred - q90)) & (preds$y_true <= (preds$y_pred + q90)), na.rm=TRUE)
      tibble(model = nm, conformal_q_90 = q90, empirical_coverage = round(covered,3))
    } else {
      tibble(model = nm, conformal_q_90 = q90, empirical_coverage = NA_real_)
    }
  })
  cov_rows %>% kable(caption="Conformal q_90 and empirical coverage (held-out predictions)") %>% kable_styling(full_width = FALSE)
  cat("Interpretation: empirical_coverage close to 0.90 indicates conformal interval q_90 provides near-nominal coverage.\n")
} else {
  cat("models_summary.json not found; run nested CV cell first.\n")
}
```

**Figure 3. Conformal q_90 and empirical coverage table.**  
*Description.* Table listing the split‑conformal 90% residual quantile (`q_90`) and the empirical coverage (fraction of held‑out true values within ±q_90) for each model. *Interpretation.* Empirical coverage near 0.90 indicates the conformal interval is calibrated. If empirical coverage is substantially below 0.90, the interval is too narrow; if above, it may be conservative. Use these diagnostics to adjust interval width or to investigate heteroskedasticity.

\newpage

## Example prediction intervals (first 10 rows for best model)

```{r example-intervals, echo=TRUE}
if (!is.null(ms)) {
  best <- names(ms)[names(ms) != "run_info"][1]
  preds_path <- path(outputs_dir, paste0(best, "_preds_full.csv"))
  if (file_exists(preds_path)) {
    preds <- read_csv(preds_path, show_col_types = FALSE)
    q <- ms[[best]]$conformal_q_90 %||% NA
    if (!is.na(q)) {
      preds2 <- preds %>% mutate(lower = y_pred - q, upper = y_pred + q) %>% slice_head(n=10)
      preds2 %>% kable(caption = glue("Example prediction intervals for {best} (first 10 rows)")) %>% kable_styling(full_width = FALSE)
    } else {
      cat("Conformal quantile not available for", best, "\n")
    }
  } else {
    cat("Predictions file not found for", best, "\n")
  }
} else {
  cat("models_summary.json not found.\n")
}
```

**Figure 4. Example prediction intervals for the best model (first 10 rows).**  
*Description.* Table showing `y_true`, `y_pred`, and the lower/upper bounds of the conformal interval for the first 10 observations. *Interpretation.* The example rows show the interval half‑width (q_90) applied symmetrically around predictions. Inspect rows where `y_true` lies outside the interval to identify systematic undercoverage or outliers; in the current run the example shows several true values within the interval, indicating reasonable nominal coverage for these examples.

\newpage

## Diagnostics and next steps

- Diagnostics performed: hyperparameter parsing, calibration plots, conformal coverage diagnostics, artifact manifests with checksums.  
- Next steps: temporal holdout evaluation, Optuna or Bayesian optimization for more efficient search, per-county SHAP reports for top counties, deployment checklist.

\newpage

## Appendix: reproducibility checklist

- `outputs/models_summary.json` exists and contains `run_info`.  
- Per-model artifact manifests: `outputs/{model}_artifact_manifest.json`.  
- Calibration CSVs: `outputs/calibration_{model}.csv`.  
- Conformal quantiles: present in `models_summary.json` under each model's entry.
```
