---
title: "County Childhood Vaccination Uptake: Final Report"
author: "Geoffrey Bruder"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  pdf_document:
    latex_engine: xelatex
    toc: true
    number_sections: true
bibliography: references.bib
csl: apa.csl
fontsize: 12pt
geometry: margin=1in
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE, fig.width = 6.5, fig.height = 4)
library(tidyverse); library(glue); library(jsonlite); library(kableExtra); library(here)

proj_root <- here::here()
outputs_dir <- file.path(proj_root, "outputs")
fig_dir <- file.path(proj_root, "figures")

# Ensure apa.csl and references.bib exist (download/create minimal if missing)
ensure_csl_and_bib <- function(csl_path = file.path(proj_root, "apa.csl"),
                               bib_path = file.path(proj_root, "references.bib")) {
  if (!file.exists(csl_path)) {
    tryCatch({
      utils::download.file("https://raw.githubusercontent.com/citation-style-language/styles/master/apa.csl",
                           destfile = csl_path, quiet = TRUE, mode = "wb")
    }, error = function(e) { })
  }
  if (!file.exists(bib_path)) {
    bib_text <- '
@misc{cdc_vaxview,
  title = {CDC VaxView County Vaccination Data},
  author = {{Centers for Disease Control and Prevention}},
  year = {2021},
  url = {https://www.cdc.gov/vaccines/vaxview/index.html},
  note = {Accessed 2026-04-25}
}
@misc{acs_2021,
  title = {American Community Survey 2021 5-Year Estimates},
  author = {{U.S. Census Bureau}},
  year = {2021},
  url = {https://www.census.gov/programs-surveys/acs},
  note = {Accessed 2026-04-25}
}
@misc{covidcast,
  title = {COVIDcast: Facebook Survey Signals},
  author = {{Delphi Research Group, Carnegie Mellon University}},
  year = {2021},
  url = {https://delphi.cmu.edu/covidcast/},
  note = {Accessed 2026-04-25}
}
'
    writeLines(text = bib_text, con = bib_path)
  }
}
ensure_csl_and_bib()

# Helper: escape LaTeX special characters (use for captions and inline text)
escape_latex <- function(x) {
  if (is.null(x)) return(x)
  x <- as.character(x)
  x <- gsub("\\\\", "\\\\textbackslash{}", x)   # backslash first
  x <- gsub("([%$#&_{}~^])", "\\\\\\1", x, perl = TRUE)
  x
}

# Helper: safe include graphics
include_fig <- function(fname, caption = NULL) {
  f <- file.path(fig_dir, fname)
  if (!file.exists(f)) {
    cat(sprintf("Figure '%s' not found. Run notebook to generate and place in figures/.\n", fname))
  } else {
    if (!is.null(caption)) {
      caption <- escape_latex(caption)
      knitr::include_graphics(f)
    } else {
      knitr::include_graphics(f)
    }
  }
}
```

# Abstract

This project develops and evaluates predictive models for county-level childhood vaccination coverage using merged CDC and American Community Survey (ACS) data, with optional Facebook COVIDcast survey signals. The analysis emphasizes reproducible preprocessing, robust imputation, nested cross-validation for model selection, and uncertainty quantification via conformal intervals. ElasticNet emerged as a competitive, interpretable baseline (CV RMSE ≈ 4.07). Diagnostics reveal heteroskedastic residuals and occasional large errors, motivating recommendations for additional data collection, temporal holdouts, and heteroskedastic-aware uncertainty quantification. The report documents methods, results, limitations, and actionable recommendations for public‑health stakeholders. (Keywords: vaccination coverage, ElasticNet, nested cross-validation, conformal prediction, reproducibility)

#  Introduction

##  Research question
Which county-level demographic, socioeconomic, and survey-derived features predict childhood vaccination coverage, and can a reproducible model identify counties at risk of low uptake to inform targeted interventions?

##  Motivation and significance
Accurate county-level estimates enable efficient allocation of public-health resources. Predictive models can identify counties likely to have low vaccination coverage, supporting targeted outreach and policy decisions.

#  Data and preprocessing

##  Data sources and acquisition
Data sources used in this analysis:

- **CDC VaxView**: county-level vaccination coverage estimates (Centers for Disease Control and Prevention).  
- **American Community Survey (ACS) 2021**: county-level demographic and socioeconomic covariates (U.S. Census Bureau).  
- **COVIDcast (optional)**: Facebook survey-derived `vaccinate_children` signal (Delphi Research Group, CMU).

(Full citations in References.)

##  Ingestion and provenance
The ingestion step pads FIPS codes, merges datasets on `fips`, coerces numeric columns, and writes `data_clean/analysis_table.csv`. Provenance is recorded in `outputs/run_provenance.json` (timestamp and environment snapshot).

##  Cleaning and imputation
- Numeric features: median imputation (global) with optional group (state) median imputation.  
- Missingness indicators appended for numeric features.  
- Categorical features: most-frequent imputation and one-hot encoding.  
- Preprocessor saved as `models/preprocessor.joblib`; numeric imputer saved as `models/simple_imputer_numeric_median.joblib`.

**Imputation audit (notebook output):** original_rows = 1588; kept_rows = 1543; dropped_rows = 45.

#  Exploratory data analysis

##  Descriptive statistics and distributions

```{r desc-stats, echo=FALSE}
data_clean_path <- file.path(proj_root, "data_clean", "analysis_table.csv")
if (file.exists(data_clean_path)) {
  df <- readr::read_csv(data_clean_path, show_col_types = FALSE)
  summary_tbl <- df %>% summarise(
    n = n(),
    mean_coverage = mean(coverage_estimate, na.rm = TRUE),
    sd_coverage = sd(coverage_estimate, na.rm = TRUE),
    min_coverage = min(coverage_estimate, na.rm = TRUE),
    max_coverage = max(coverage_estimate, na.rm = TRUE)
  ) %>% pivot_longer(everything(), names_to = "metric", values_to = "value")
  caption_text <- escape_latex("Descriptive summary of target (coverage_estimate)")
  summary_tbl %>% knitr::kable(caption = caption_text) %>% kable_styling(full_width = FALSE)
} else {
  cat("analysis_table.csv not found in data_clean/. Run ingestion cell to create it.\n")
}
```

##  Missingness patterns (figure)

**Figure 1. Per-state missingness for representative features.**  
*Caption.* Counts of missing values filled by group imputation per state for top features. *Reference in text.* The per‑state missingness plot (Figure 1) shows `vaccinate_children` has substantial missingness in many counties; group imputation filled many ACS variables using state medians.

```{r fig-missing, echo=FALSE, fig.cap=escape_latex("Figure 1. Per-state missingness for representative features (group imputation audit).")}
f <- file.path(fig_dir, "per_state_missing.png")
if (file.exists(f)) knitr::include_graphics(f) else cat("Figure 'per_state_missing.png' not found; run notebook to generate.\n")
```

#  Modeling approach

##  Preprocessing pipeline
Preprocessor: median imputer → StandardScaler for numeric; most-frequent imputer → OneHotEncoder for categorical. Missingness indicators retained.

##  Candidate algorithms
- ElasticNet (regularized linear model) — interpretable baseline.  
- Random Forest — ensemble tree model.  
- Gradient Boosting (sklearn/LightGBM) — boosted trees.  
- MLP (PyTorch) — neural network baseline.

##  Validation and tuning
Nested cross-validation: outer 5-fold for performance estimation; inner 3-fold RandomizedSearchCV for hyperparameter search. Final refit on full cleaned dataset with best hyperparameters saved.

#  Results

##  Nested CV summary (table)

```{r cv-summary, echo=FALSE}
ms_path <- file.path(outputs_dir, "models_summary.json")
if (file.exists(ms_path)) {
  ms <- jsonlite::fromJSON(ms_path)
  models <- ms[names(ms) != "run_info"]
  df_models <- map_dfr(names(models), function(nm) {
    v <- models[[nm]]
    tibble(
      model = nm,
      cv_rmse_mean = v$cv_rmse_mean %||% NA,
      cv_rmse_std = v$cv_rmse_std %||% NA,
      cv_mae_mean = v$cv_mae_mean %||% NA,
      cv_r2_mean = v$cv_r2_mean %||% NA
    )
  })
  df_models %>% arrange(cv_rmse_mean) %>% mutate(across(where(is.numeric), ~ round(.x,3))) %>% knitr::kable(caption = escape_latex("Table 1. Nested CV model comparison (lower RMSE better)")) %>% kable_styling(full_width = FALSE)
} else {
  cat("models_summary.json not found in outputs/. Run the notebook to generate nested CV results.\n")
}
```

**Interpretation.** ElasticNet CV RMSE ≈ 4.07 (best by mean RMSE); Random Forest and GBM similar; MLP underperformed.

##  Residual diagnostics (figure)

**Figure 2. Residuals vs predicted and residual histogram (ElasticNet).**  
*Caption.* Residuals vs predicted values and residual histogram for the ElasticNet final model. These diagnostics show residuals centered near zero but with heavier tails and heteroskedasticity.

```{r fig-resid, echo=FALSE, fig.cap=escape_latex("Figure 2. Residual diagnostics for ElasticNet (residuals vs predicted; histogram).")}
f1 <- file.path(fig_dir, "residuals_vs_predicted.png")
f2 <- file.path(fig_dir, "residual_histogram.png")
if (file.exists(f1)) knitr::include_graphics(f1) else cat("Figure 'residuals_vs_predicted.png' not found.\n")
if (file.exists(f2)) knitr::include_graphics(f2) else cat("Figure 'residual_histogram.png' not found.\n")
```

##  Conformal intervals and calibration (table + figure)

**Table 2.** Conformal q_90 and empirical coverage per model (if available in `models_summary.json`).

```{r conformal-table, echo=FALSE}
if (file.exists(ms_path)) {
  cov_rows <- map_dfr(names(ms)[names(ms) != "run_info"], function(nm) {
    v <- ms[[nm]]
    q90 <- v$conformal_q_90 %||% NA
    preds_path <- file.path(outputs_dir, paste0(nm, "_preds_full.csv"))
    if (file.exists(preds_path) && !is.na(q90)) {
      preds <- readr::read_csv(preds_path, show_col_types = FALSE)
      covered <- mean((preds$y_true >= (preds$y_pred - q90)) & (preds$y_true <= (preds$y_pred + q90)), na.rm = TRUE)
      tibble(model = nm, conformal_q_90 = q90, empirical_coverage = round(covered, 3))
    } else {
      tibble(model = nm, conformal_q_90 = q90, empirical_coverage = NA_real_)
    }
  })
  cov_rows %>% knitr::kable(caption = escape_latex("Table 2. Conformal q_90 and empirical coverage (held-out predictions)")) %>% kable_styling(full_width = FALSE)
} else {
  cat("models_summary.json not found; run nested CV cell first.\n")
}
```

#  Interpretation and stakeholder implications

ElasticNet coefficients (see `outputs/elasticnet_coefficients.csv`) identify features associated with higher or lower coverage; positive coefficients indicate features associated with higher predicted coverage, negative with lower. Use coefficient directionality and SHAP summaries to craft targeted interventions (e.g., counties with low income and low education indicators may be prioritized for outreach).

#  Limitations

- Missingness in survey-derived features reduces predictive power.  
- Cross-sectional evaluation only; temporal generalization untested.  
- Heteroskedastic residuals suggest variance-stabilizing transforms or heteroskedastic-aware intervals may improve uncertainty quantification.

#  Recommendations and next steps

1. Collect additional local survey data to reduce missingness in `vaccinate_children`.  
2. Run temporal holdout evaluation before deployment.  
3. Implement heteroskedastic-aware conformal intervals or model residual variance explicitly.  
4. Produce per-county SHAP one‑page summaries for stakeholder use.

#  Methods appendix

##  Reproducible run instructions

1. Place raw files in `data_raw/`: CDC, ACS, optional COVIDcast.  
2. Run `notebooks/analysis_updated_full.ipynb` top→bottom.  
3. Knit this Rmd with `rmarkdown::render("M08_Final_Written_Deliverables.Rmd")`.

##  Ethical considerations and data use agreements

All data used are publicly available aggregate county-level data. No individual-level or personally identifiable information (PII) is used. Users must comply with the terms of use for each data source (CDC, ACS, COVIDcast). When using survey-derived signals (COVIDcast), ensure appropriate citation and respect any data-use restrictions; do not attempt to re-identify individuals from aggregated signals. For deployment, consider equity implications of model-driven interventions and consult local public-health authorities before acting on model outputs.

#  Appendix

##  Data dictionary and artifacts

- `outputs/data_dictionary.csv` — auto-generated column types and sample values.  
- `outputs/data_imputation_and_cleaning_summary.json` — imputation audit.  
- `outputs/group_imputation_audit.json` — group imputation audit.  
- `outputs/models_summary.json` — nested CV and model artifacts.

## Figure legends (APA-style)

**Figure 1.** Per-state missingness for representative features (group imputation audit). Legend: bars show counts of missing values filled by group imputation per state for selected features; `vaccinate_children` indicates survey-derived missingness.

**Figure 2.** Residual diagnostics for ElasticNet. Legend: left panel shows residuals vs predicted values with horizontal line at zero; right panel shows residual histogram with bin width chosen to reveal tail behavior.

## 10.3 Unit tests and testing summary

A minimal `tests/` example (R `testthat`) is recommended. Example file `tests/test_core.R`:

```r
# tests/test_core.R
library(testthat)
test_that("checksum_file returns hex string of length 64 for existing file", {
  f <- "README.md"
  if (file.exists(f)) {
    ch <- tools::md5sum(f)
    expect_true(nchar(as.character(ch)) > 0)
  } else {
    skip("README.md not present in repo root; skip checksum test.")
  }
})
```

**Testing summary.** Add `tests/` to repo and run:

```r
# in R
install.packages("testthat")
testthat::test_dir("tests")
```

# References

(APA 7 formatted references will render from `references.bib`.)
```