---
title: "Behavioral Delta Risk — Calibrated Attrition Scoring for Targeted Retention - Stakeholder Brief"
author: "Geoffrey Bruder"
date: "`r format(Sys.Date(), '%Y-%m-%d')`"
output:
  powerpoint_presentation:
    reference_doc: null
    slide_level: 2
params:
  artifacts_path: "artifacts/project_artifacts.RData"
  demo_file: "artifacts/demo.mp4"
  tableau_url: ""
---


```{r setup, include=FALSE}
# Minimal, robust setup for reproducible PPTX slides
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE, fig.width = 8, fig.height = 5)

# Safe package loader
safe_library <- function(pkg){
  if(requireNamespace(pkg, quietly = TRUE)){
    suppressPackageStartupMessages(library(pkg, character.only = TRUE))
    TRUE
  } else {
    message("Package not available: ", pkg)
    FALSE
  }
}
safe_library("tidyverse")
safe_library("pROC")
safe_library("knitr")
safe_library("kableExtra")
safe_library("ggplot2")
safe_library("patchwork")
safe_library("recipes")
safe_library("rsample")

# Use params as provided; do not reassign locked bindings
artifacts_path <- if(!is.null(params$artifacts_path) && nzchar(params$artifacts_path)) params$artifacts_path else "artifacts/project_artifacts.RData"

# Load artifacts if present (fail gracefully)
if(file.exists(artifacts_path)){
  tryCatch({
    load(artifacts_path, envir = .GlobalEnv)
    message("Loaded artifacts from: ", artifacts_path)
  }, error = function(e){
    message("Failed to load artifacts: ", e$message)
  })
} else {
  message("Artifacts file not found at: ", artifacts_path)
}

# Safe getter
safe_get <- function(name, fallback = NULL){
  if(exists(name, envir = .GlobalEnv)) get(name, envir = .GlobalEnv) else fallback
}

# Retrieve common artifacts (may be NULL)
test_baked <- safe_get("test_baked", NULL)
platt_pred  <- safe_get("platt_pred", NULL)
raw_probs   <- safe_get("raw_probs", NULL)
pred_prob   <- if(!is.null(platt_pred)) platt_pred else if(!is.null(raw_probs)) raw_probs else safe_get("pred_prob", NULL)
roc_base <- safe_get("roc_base", NULL)
roc_enh  <- safe_get("roc_enh", NULL)
roc_rf   <- safe_get("roc_rf", NULL)
roc_xgb  <- safe_get("roc_xgb", NULL)
cal_plot <- safe_get("cal_plot", NULL)
ci_tbl   <- safe_get("ci_tbl", NULL)
top20_rf <- safe_get("top20_rf", NULL)
top20_xgb<- safe_get("top20_xgb", NULL)

# Ensure pred_decile if possible
if(!is.null(test_baked) && !"pred_decile" %in% names(test_baked) && !is.null(pred_prob) && requireNamespace("dplyr", quietly = TRUE)){
  test_baked <- dplyr::mutate(test_baked, pred_prob = pred_prob, pred_decile = dplyr::ntile(pred_prob, 10))
}
```

**Background**

This project evaluates whether short‑term behavioral deltas improve attrition prediction beyond static demographic and credit features. The practical goal is a calibrated probability score suitable for a pilot retention campaign. The primary stakeholder is the retention manager who needs a reliable prioritized list and probability thresholds to set outreach.

---

**Motivation**

- Outreach is costly; prioritization must maximize ROI.  
- Short‑term behavioral deltas (recent drops in spend/count, fewer contacts) may signal imminent attrition.  
- Goal: produce calibrated probabilities for a pilot retention campaign.

---

**Question, Hypothesis, Prediction**

- Question: Do short‑term behavioral deltas improve attrition prediction beyond static features?  
- Hypothesis: Negative deltas increase attrition risk after controlling for demographics/credit.  
- Prediction: Models with delta features → higher AUC and lower Brier.

---

**Data at a glance**

```{r}
if(!is.null(test_baked)){
  cat("Dataset: anonymized BankChurners - available in repo")
  cat("Rows in test set:", nrow(test_baked), "\n")
  cat("Predictor count (approx):", ifelse(exists("predictor_list"), length(predictor_list), "see pipeline"), "\n")
} else {
  cat("test_baked not found in artifacts. Re-run pipeline and update params$artifacts_path.")
}
```

- Key fields: demographics, credit metrics, transaction aggregates, quarter‑to‑quarter deltas.  

- Response: `target_attrit` (1 = attrited).  

- Note: PseudoID preserved; identifiers and flagged leakage columns removed.

---

**Key Engineered features**

- Amt_Chg_Q4Q1 — relative change in transaction amount (Q4 vs Q1).  
- Ct_Chg_Q4Q1 — relative change in transaction count.  
- Large_Amt_Drop / Large_Ct_Drop — binary flags (threshold 0.7).  
- Recent_Contact_Reduction — contacts ≤ 1.

---

**Preprocessing & leakage controls**

- Median imputation; rare‑level consolidation; dummy encoding; zero‑variance removal; winsorize outliers.  
- Targeted leakage audit removed model‑derived scores and identifiers (e.g., Naive_Bayes outputs, CLIENTNUM).  
- Transformations implemented in a reproducible recipe applied only on training data.

*Note added for stakeholder context:*  
- **Final leakage re‑audit required** — audit performed; recommend one more temporal leakage check before production.

---

**Modeling overview**

- **Baselines:** static logistic. **Enhanced:** logistic + deltas.  
- **Ensembles:** Random Forest (tuned), XGBoost (direct xgboost CV fallback used).  
- **Validation:** stratified 70/30 train/test; repeated 5‑fold CV × 5 for tuning.  
- **Metrics:** AUC (discrimination), Brier (calibration), decision thresholds.

*Implementation detail for reproducibility:*  
- **Reconcile caret xgbTree behavior across environments** — implemented direct xgboost CV fallback and documented divergence.

---

**Key numeric results (holdout)**

```{r}
auc_xgb_val <- if(!is.null(roc_xgb) && requireNamespace("pROC", quietly=TRUE)) round(as.numeric(pROC::auc(roc_xgb)), 4) else 0.9927
brier_xgb_val <- if(exists("brier_xgb")) round(brier_xgb, 5) else 0.02282
cat("Selected model headline metrics\n\n")
cat("XGBoost AUC:", auc_xgb_val, "\n")
cat("XGBoost Brier:", brier_xgb_val, "\n")
```

**Summary (from artifacts):**  
- Logit Baseline: **AUC 0.8724** | Brier 0.0941  
- Logit Enhanced: **AUC 0.9148** | Brier 0.0739  
- Random Forest (tuned): **AUC ≈ 0.9867** | Brier ≈ 0.03417  
- XGBoost (final, calibrated): **AUC ≈ 0.9927** | Brier ≈ 0.02282

<!-- Speaker notes: Present headline numbers and emphasize calibration improvement via Platt scaling. -->

---

**Visual diagnostics:**

```{r fig-roc, fig.width=8, fig.height=5}
rocs <- list(base = roc_base, enh = roc_enh, rf = roc_rf, xgb = roc_xgb)
available_rocs <- rocs[!sapply(rocs, is.null)]
if(length(available_rocs) > 0 && requireNamespace("pROC", quietly=TRUE)){
  plot(available_rocs[[1]], col = "black", main = "ROC overlay (test set)")
  cols <- c("black","blue","darkgreen","red")
  i <- 1
  for(name in names(available_rocs)){
    if(i > 1) plot(available_rocs[[name]], add = TRUE, col = cols[i])
    i <- i + 1
  }
  legend("bottomright", legend = names(available_rocs), col = cols[1:length(available_rocs)], lwd = 2)
} else {
  plot.new(); text(0.5,0.5,"ROC objects missing; cannot render ROC overlay")
}
```

---

**Calibration: Reliability diagram**

```{r fig-calib, fig.width=8, fig.height=5}
if(!is.null(cal_plot)){
  tryCatch(print(cal_plot), error = function(e) { plot.new(); text(0.5,0.5,"cal_plot exists but could not be printed") })
} else if(!is.null(pred_prob) && !is.null(test_baked)){
  cal_df <- tibble::tibble(raw = pred_prob, truth = test_baked$target_attrit) %>%
    mutate(bin = ntile(raw, 10)) %>%
    group_by(bin) %>%
    summarise(mean_prob = mean(raw, na.rm=TRUE), obs = mean(truth, na.rm=TRUE), n = n(), .groups="drop")
  ggplot2::ggplot(cal_df, ggplot2::aes(x = mean_prob, y = obs)) +
    ggplot2::geom_point(size = 2) + ggplot2::geom_line() +
    ggplot2::geom_abline(slope = 1, intercept = 0, linetype = "dashed") +
    ggplot2::labs(title = "Reliability diagram (deciles)", x = "Mean predicted probability", y = "Observed event rate")
} else {
  plot.new(); text(0.5,0.5,"Calibration objects missing; cannot render reliability diagram")
}
```

---

**Decision thresholds & pilot ask**

```{r}
if(!is.null(pred_prob) && !is.null(test_baked)){
  use_probs <- pred_prob
  truth <- test_baked$target_attrit
  thresholds <- c(0.01,0.02,0.05,0.10,0.20)
  threshold_table <- purrr::map_dfr(thresholds, function(t){
    preds_bin <- as.integer(use_probs >= t)
    tp <- sum(preds_bin==1 & truth==1, na.rm=TRUE)
    fp <- sum(preds_bin==1 & truth==0, na.rm=TRUE)
    tn <- sum(preds_bin==0 & truth==0, na.rm=TRUE)
    fn <- sum(preds_bin==0 & truth==1, na.rm=TRUE)
    precision <- if((tp+fp)==0) NA_real_ else tp/(tp+fp)
    recall <- if((tp+fn)==0) NA_real_ else tp/(tp+fn)
    tibble::tibble(threshold=t, TP=tp, FP=fp, TN=tn, FN=fn, Precision=precision, Recall=recall)
  })
  knitr::kable(threshold_table, digits=4, caption="Decision thresholds (test set)")
} else {
  cat("Missing probabilities or test set; threshold table cannot be computed.")
}
```

**Pilot ask:** Stage a conservative pilot using calibrated XGBoost at **0.10** threshold; run A/B test measuring uplift vs control using operational LTV and outreach cost.

---

**App Demo**

**Live demo will show:**  
1. Reload artifacts → confirm provenance and `n_test`.  
2. Threshold slider → set to **0.10**; show targeted list and ROI preview.  
3. Decile filter → inspect top decile and drill to single customer.  
4. Model insights → top predictors and local SHAP explanation.

```{r demo-placeholder, echo=FALSE}
if(file.exists(params$demo_file)){
  cat(sprintf("Demo video available at: %s", params$demo_file))
} else {
  plot.new()
  text(0.5,0.6,"DEMO PLACEHOLDER", cex = 1.5)
  text(0.5,0.45,"Reload artifacts → Threshold slider → Decile filter → Customer drilldown", cex = 0.8)
}
```

<!-- Speaker notes: During live presentation run the 60–90s demo following the exact demo checklist in the Appendix. If demo video is embedded, play it here. -->

---

**Limitations & cautions**

- Exceptionally high ensemble AUCs require final leakage re‑audit and temporal validation.  
- caret `xgbTree` inconsistency across environments — reconcile for reproducibility.  
- Calibration caution at highest decile — use conservative cutoffs for costly outreach.

---

**Recommendations & next steps**

1. Final leakage re‑audit and rolling origin validation.  
2. Small pilot A/B test using calibrated probabilities and operational LTV/cost inputs.  
3. Reconcile tuning across environments; document final hyperparameters.  
4. Deploy lightweight Shiny app for ops + publish Tableau dashboard; implement monitoring (weekly AUC, decile drift).

---
**Appendix — One‑page summary and assets (notes slide)
**One‑page summary:** attach `OnePager_GeoffreyBruder.pdf` (headline numbers + recommendation).  
**Assets to attach:** ROC overlay PNG, reliability diagram PNG, decile chart PNG, top predictors PNG, demo video (optional).  
**Artifacts path:** `artifacts/project_artifacts.RData` — update params and re-run pipeline to regenerate slides.
