---
title: "Behavioral Delta Risk: Calibrated Attrition Scoring for Targeted Retention_PRIOR TO TUNING"
author: "Geoffrey Bruder"
date: "`r format(Sys.Date(), '%Y-%m-%d')`"
output:
  pdf_document:
    latex_engine: xelatex
    toc: true
    toc_depth: 4
  html_document:
    toc: true
    toc_depth: 2
    toc_float: true
editor_options:
  markdown:
    wrap: 72
header-includes:
  - \usepackage{pdflscape}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
set.seed(2025)
```

# 0. Introduction

Behavioral Delta Risk: Calibrated Attrition Scoring for Targeted Retention, develops a predictive model that estimates each customer’s risk of attrition conditional on recent behavioral change, specifically the probability of attrition given a recent drop or spike in transaction volume or product interactions. The primary stakeholder for this work is the Retention Marketing and Customer Lifecycle Management team at ABC Corporation, who require calibrated, actionable risk scores to prioritize outreach. My central hypothesis is that customers exhibiting recent negative behavioral changes—notably large declines in transaction amount or transaction count and reductions in contact frequency—are more likely to attrite after controlling for demographic and credit metrics. I predict that models incorporating engineered behavioral‑change features (continuous delta measures and binary drop flags) will outperform static‑only models on discrimination (AUC) and calibration (Brier score and reliability), and that these gains will be sufficient to support operational thresholding for targeted retention campaigns.

---

# 1. Data import and initial cleaning

```{r packages, echo=TRUE}
# Required packages
packages <- c(
  "tidyverse","caret","kableExtra","pROC","gridExtra","glmnet",
  "randomForest","xgboost","vip","scales","recipes","rsample","yardstick","themis","broom"
)
installed <- rownames(installed.packages())
for(p in packages) if(!(p %in% installed)) install.packages(p, dependencies = TRUE)
invisible(lapply(packages, library, character.only = TRUE))
```

```{r load-data}
library(here)

# resolve path relative to project root
data_path <- here("data_raw", "BankChurners.csv")

# clear, actionable error if missing
if (!file.exists(data_path)) {
  stop("data_raw/BankChurners.csv not found. Place BankChurners.csv in the project data_raw folder (", data_path, ").")
}

# read into the object your code expects
customer_data <- read.csv(data_path, stringsAsFactors = FALSE)


# Create PseudoID if not present
if(!"PseudoID" %in% names(customer_data)){
  customer_data <- customer_data %>% mutate(PseudoID = paste0("CUST", row_number()))
}

# Extract CLIENTNUM mapping if present and remove from analysis
if("CLIENTNUM" %in% names(customer_data)){
  client_map <- customer_data %>% select(PseudoID, CLIENTNUM)
  customer_data <- customer_data %>% select(-CLIENTNUM)
} else {
  client_map <- NULL
}

message("Loaded BankChurners.csv: ", nrow(customer_data), " rows, ", ncol(customer_data), " columns")
glimpse(customer_data)
```

---

# 2. Feature engineering and preprocessing

```{r preprocess}
# Standardize column names
names(customer_data) <- make.names(names(customer_data))
data_full <- customer_data

# Categorical variables (if present)
possible_cat <- c("Attrition_Flag","Gender","Education_Level","Marital_Status","Income_Category","Card_Category")
cat_vars <- intersect(possible_cat, names(data_full))
if(length(cat_vars) > 0){
  data_full[cat_vars] <- lapply(data_full[cat_vars], function(x) ifelse(is.na(x) | x=="", "Unknown", x))
  data_full[cat_vars] <- lapply(data_full[cat_vars], factor)
}

# Numeric variables (if present)
possible_num <- c(
  "Customer_Age","Dependent_count","Months_on_book","Total_Relationship_Count",
  "Months_Inactive_12_mon","Contacts_Count_12_mon","Credit_Limit","Total_Revolving_Bal",
  "Avg_Open_To_Buy","Total_Amt_Chng_Q4_Q1","Total_Trans_Amt","Total_Trans_Ct",
  "Total_Ct_Chng_Q4_Q1","Avg_Utilization_Ratio"
)
num_vars <- intersect(possible_num, names(data_full))
data_full[num_vars] <- lapply(data_full[num_vars], function(x) as.numeric(as.character(x)))

# Behavioral-change features (novelty)
data_full <- data_full %>%
  mutate(
    Amt_Chg_Q4Q1 = ifelse("Total_Amt_Chng_Q4_Q1" %in% names(.), Total_Amt_Chng_Q4_Q1, NA_real_),
    Ct_Chg_Q4Q1  = ifelse("Total_Ct_Chng_Q4_Q1" %in% names(.), Total_Ct_Chng_Q4_Q1, NA_real_),
    Large_Amt_Drop = ifelse(!is.na(Amt_Chg_Q4Q1) & Amt_Chg_Q4Q1 < 0.7, 1, 0),
    Large_Ct_Drop  = ifelse(!is.na(Ct_Chg_Q4Q1) & Ct_Chg_Q4Q1 < 0.7, 1, 0),
    Recent_Contact_Reduction = ifelse(!is.na(Contacts_Count_12_mon) & Contacts_Count_12_mon <= 1, 1, 0)
  )

# Binary target
if("Attrition_Flag" %in% names(data_full)){
  data_full <- data_full %>%
    mutate(target_attrit = case_when(
      grepl("Attrited", Attrition_Flag, ignore.case = TRUE) ~ 1,
      grepl("Existing", Attrition_Flag, ignore.case = TRUE) ~ 0,
      TRUE ~ NA_real_
    ))
} else {
  data_full$target_attrit <- NA_real_
}

# Modeling subset (rows with known target)
model_data <- data_full %>% filter(!is.na(target_attrit))
message("Rows available for modeling: ", nrow(model_data))
```

---

# 3. Exploratory checks

```{r eda}
# Quick frequency for Attrition_Flag if present
if("Attrition_Flag" %in% names(model_data)){
  knitr::kable(as.data.frame(table(model_data$Attrition_Flag)), caption = "Attrition_Flag counts")
}

# Numeric summary
if(ncol(select_if(model_data, is.numeric)) > 0){
  print(summary(select_if(model_data, is.numeric)))
}
```

---

# 4. Modeling plan and reproducible pipeline

**Plan:** split 70/30 stratified; baseline logistic (static), enhanced logistic (adds behavioral-change), Random Forest, XGBoost (direct xgboost workflow); compare AUC, calibration; export scores.

```{r split-and-recipe}
set.seed(2025)
split <- initial_split(model_data, prop = 0.7, strata = "target_attrit")
train <- training(split)
test  <- testing(split)

# Build recipe with the predictors that exist
predictors <- c(
  intersect("Customer_Age", names(train)),
  intersect("Gender", names(train)),
  intersect("Credit_Limit", names(train)),
  intersect("Total_Trans_Amt", names(train)),
  intersect("Total_Trans_Ct", names(train)),
  intersect("Avg_Utilization_Ratio", names(train)),
  intersect("Total_Relationship_Count", names(train)),
  intersect("Months_on_book", names(train)),
  intersect("Amt_Chg_Q4Q1", names(train)),
  intersect("Ct_Chg_Q4Q1", names(train)),
  intersect("Large_Amt_Drop", names(train)),
  intersect("Large_Ct_Drop", names(train)),
  intersect("Recent_Contact_Reduction", names(train))
)
predictors <- predictors[predictors != ""]

rec <- recipe(as.formula(paste("target_attrit ~", paste(predictors, collapse = " + "))), data = train) %>%
  step_impute_median(all_numeric(), -all_outcomes()) %>%
  step_unknown(all_nominal(), -all_outcomes()) %>%
  step_other(all_nominal(), -all_outcomes(), threshold = 0.01) %>%
  step_dummy(all_nominal(), -all_outcomes()) %>%
  step_zv(all_predictors()) %>%
  step_normalize(all_numeric(), -all_outcomes())

prep_rec <- prep(rec, training = train)
train_baked <- bake(prep_rec, new_data = train)
test_baked  <- bake(prep_rec, new_data = test)

# caret expects factor outcome with levels "Yes"/"No"
train_baked <- train_baked %>% mutate(target = factor(ifelse(target_attrit==1, "Yes", "No")))
test_baked  <- test_baked  %>% mutate(target = factor(ifelse(target_attrit==1, "Yes", "No")))

# For caret x/y interface: predictors only (remove target_attrit and target)
x_train <- train_baked %>% select(-target_attrit, -target)
y_train <- train_baked$target
x_test  <- test_baked %>% select(-target_attrit, -target)
y_test  <- test_baked$target

# Defensive conversions: ensure all predictor columns are numeric
x_train <- x_train %>% mutate(across(everything(), ~ as.numeric(as.character(.))))
x_test  <- x_test  %>% mutate(across(everything(), ~ as.numeric(as.character(.))))

# Check for NA/Inf in predictors; impute with column medians if any remain
na_counts_train <- colSums(is.na(x_train) | is.infinite(as.matrix(x_train)))
if(any(na_counts_train > 0)){
  message("Found NA/Inf in training predictors; imputing with column medians for these columns.")
  for(col in names(x_train)[na_counts_train > 0]){
    med <- median(x_train[[col]], na.rm = TRUE)
    x_train[[col]][is.na(x_train[[col]]) | is.infinite(x_train[[col]])] <- med
    x_test[[col]][is.na(x_test[[col]]) | is.infinite(x_test[[col]])] <- med
  }
}

# Final checks
if(!is.data.frame(x_train)) x_train <- as.data.frame(x_train)
if(!is.data.frame(x_test)) x_test <- as.data.frame(x_test)

# Ensure y_train has exactly two levels
if(length(levels(y_train)) != 2) stop("y_train does not have exactly two factor levels. Check target encoding.")
message("Prepared x_train (", nrow(x_train), " x ", ncol(x_train), ") and y_train with levels: ", paste(levels(y_train), collapse = ", "))
```

---

# 5. Baseline and enhanced logistic regression

```{r baseline-enhanced}
# Baseline: static features only (rebuild recipe for static subset)
static_preds <- intersect(c("Customer_Age","Gender","Credit_Limit","Total_Trans_Amt","Total_Trans_Ct","Avg_Utilization_Ratio","Total_Relationship_Count","Months_on_book"), names(train))
static_preds <- static_preds[static_preds != ""]

rec_base <- recipe(as.formula(paste("target_attrit ~", paste(static_preds, collapse = " + "))), data = train) %>%
  step_impute_median(all_numeric(), -all_outcomes()) %>%
  step_unknown(all_nominal(), -all_outcomes()) %>%
  step_dummy(all_nominal(), -all_outcomes()) %>%
  step_normalize(all_numeric(), -all_outcomes())

train_base <- bake(prep(rec_base, training = train), new_data = train)
test_base  <- bake(prep(rec_base, training = train), new_data = test)

glm_base <- glm(target_attrit ~ ., data = train_base, family = binomial)
pred_base_prob <- predict(glm_base, newdata = test_base, type = "response")
roc_base <- pROC::roc(test_base$target_attrit, pred_base_prob, quiet = TRUE)
auc_base <- pROC::auc(roc_base)

# Enhanced logistic (use baked train_baked/test_baked)
enh_predictors <- names(train_baked)[!names(train_baked) %in% c("target_attrit","target")]
train_enh_df <- train_baked %>% select(all_of(c("target_attrit", enh_predictors)))
test_enh_df  <- test_baked  %>% select(all_of(c("target_attrit", enh_predictors)))

glm_enh <- glm(target_attrit ~ ., data = train_enh_df, family = binomial)
pred_enh_prob <- predict(glm_enh, newdata = test_enh_df, type = "response")
roc_enh <- pROC::roc(test_enh_df$target_attrit, pred_enh_prob, quiet = TRUE)
auc_enh <- pROC::auc(roc_enh)

message("AUC baseline (static logit): ", round(as.numeric(auc_base), 4))
message("AUC enhanced (logit + behavioral): ", round(as.numeric(auc_enh), 4))
```

---

# 6. Random Forest (caret) — x/y interface

```{r rf-tune}
set.seed(2025)
ctrl <- trainControl(
  method = "repeatedcv", number = 5, repeats = 2,
  classProbs = TRUE, summaryFunction = twoClassSummary,
  savePredictions = "final", verboseIter = FALSE,
  allowParallel = FALSE
)

rf_grid <- expand.grid(mtry = c(3, 5, 7))

rf_model <- tryCatch({
  train(x = x_train, y = y_train,
        method = "rf",
        metric = "ROC",
        trControl = ctrl,
        tuneGrid = rf_grid,
        ntree = 500)
}, error = function(e){
  stop("Random Forest training failed: ", e$message)
})

rf_pred_prob <- predict(rf_model, newdata = x_test, type = "prob")[, "Yes"]
roc_rf <- pROC::roc(as.numeric(y_test == "Yes"), rf_pred_prob, quiet = TRUE)
auc_rf <- pROC::auc(roc_rf)
message("AUC Random Forest: ", round(as.numeric(auc_rf), 4))
```

---

# 7. XGBoost (direct xgboost workflow) — robust nrounds selection

```{r xgb-direct}
# Convert x_train/x_test to numeric matrices and labels to numeric 0/1
xgb_train_matrix <- as.matrix(x_train)
xgb_test_matrix  <- as.matrix(x_test)
label_train <- ifelse(y_train == "Yes", 1L, 0L)
label_test  <- ifelse(y_test == "Yes", 1L, 0L)

# Final safety checks
if(any(is.na(xgb_train_matrix)) || any(is.infinite(xgb_train_matrix))){
  stop("xgb_train_matrix contains NA or Inf values. Please inspect and impute.")
}
if(!is.numeric(label_train)) stop("label_train must be numeric 0/1.")

# Parameter grid to search (small)
param_grid <- expand.grid(
  max_depth = c(3, 6),
  eta = c(0.05, 0.1),
  subsample = c(0.7, 0.8),
  colsample_bytree = c(0.7, 0.8),
  min_child_weight = c(1, 3)
)

# Use xgb.cv to pick nrounds for each param combo (lightweight CV)
best_result <- NULL
best_auc <- -Inf
for(i in seq_len(nrow(param_grid))){
  params <- list(
    objective = "binary:logistic",
    eval_metric = "auc",
    max_depth = param_grid$max_depth[i],
    eta = param_grid$eta[i],
    subsample = param_grid$subsample[i],
    colsample_bytree = param_grid$colsample_bytree[i],
    min_child_weight = param_grid$min_child_weight[i],
    verbosity = 0
  )
  dtrain <- xgboost::xgb.DMatrix(data = xgb_train_matrix, label = label_train)
  cv <- tryCatch({
    xgboost::xgb.cv(
      params = params,
      data = dtrain,
      nrounds = 200,
      nfold = 5,
      showsd = TRUE,
      early_stopping_rounds = 20,
      verbose = FALSE,
      maximize = TRUE
    )
  }, error = function(e) {
    warning("xgb.cv failed for param row ", i, ": ", e$message)
    return(NULL)
  })
  if(!is.null(cv)){
    # cv may be a list with element 'evaluation_log' and possibly 'best_iteration'
    eval_log <- as.data.frame(cv$evaluation_log)
    # Determine best iteration robustly
    best_iter <- NULL
    if(!is.null(cv$best_iteration) && !is.na(cv$best_iteration) && cv$best_iteration > 0){
      best_iter <- as.integer(cv$best_iteration)
    } else if("test_auc_mean" %in% names(eval_log)){
      best_iter <- which.max(eval_log$test_auc_mean)
    } else if("test_auc" %in% names(eval_log)){
      best_iter <- which.max(eval_log$test_auc)
    }
    # Guard: ensure best_iter is valid
    if(is.null(best_iter) || !is.finite(best_iter) || best_iter < 1){
      warning("Could not determine a valid best iteration for param row ", i, "; skipping.")
      next
    }
    best_score <- max(eval_log[[grep("test_auc", names(eval_log), value = TRUE)[1]]], na.rm = TRUE)
    if(is.na(best_score)) best_score <- -Inf
    if(best_score > best_auc){
      best_auc <- best_score
      best_result <- list(params = params, nrounds = as.integer(best_iter), cv = cv)
    }
  }
}

if(is.null(best_result)){
  stop("xgboost CV failed for all parameter combinations. Inspect warnings() for details.")
}

# Validate nrounds
nrounds_final <- best_result$nrounds
if(is.null(nrounds_final) || !is.finite(nrounds_final) || nrounds_final < 1){
  message("Selected nrounds invalid; falling back to default nrounds = 50")
  nrounds_final <- 50L
}

message("Selected XGBoost params: max_depth=", best_result$params$max_depth,
        " eta=", best_result$params$eta,
        " subsample=", best_result$params$subsample,
        " colsample_bytree=", best_result$params$colsample_bytree,
        " min_child_weight=", best_result$params$min_child_weight,
        " nrounds=", nrounds_final)

# Train final xgboost model
dtrain_final <- xgboost::xgb.DMatrix(data = xgb_train_matrix, label = label_train)
xgb_final <- tryCatch({
  xgboost::xgb.train(
    params = best_result$params,
    data = dtrain_final,
    nrounds = nrounds_final,
    verbose = 0,
    watchlist = list(train = dtrain_final),
    maximize = TRUE
  )
}, error = function(e){
  stop("xgboost::xgb.train failed: ", e$message)
})

# Predict and compute AUC
dtest <- xgboost::xgb.DMatrix(data = xgb_test_matrix)
xgb_pred_prob <- predict(xgb_final, newdata = dtest)
roc_xgb <- pROC::roc(label_test, xgb_pred_prob, quiet = TRUE)
auc_xgb <- pROC::auc(roc_xgb)
message("AUC XGBoost (direct): ", round(as.numeric(auc_xgb), 4))
```

---

# 8. Model comparison and ROC plots

```{r compare}
auc_table <- tibble(
  Model = c("Logit Baseline", "Logit Enhanced", "Random Forest", "XGBoost"),
  AUC = c(as.numeric(auc_base), as.numeric(auc_enh), as.numeric(auc_rf), as.numeric(auc_xgb))
)
knitr::kable(auc_table, caption = "Model AUC comparison (test set)")
```

```{r plot-rocs, fig.width=7, fig.height=6}
plot(roc_base, col = "black", main = "ROC curves (test set)")
plot(roc_enh, add = TRUE, col = "blue")
plot(roc_rf, add = TRUE, col = "darkgreen")
plot(roc_xgb, add = TRUE, col = "red")
legend("bottomright", legend = c("Logit Baseline","Logit Enhanced","Random Forest","XGBoost"),
       col = c("black","blue","darkgreen","red"), lwd = 2)
```

---

# 9. Calibration and interpretability

```{r calibration}
# Choose best model by AUC
aucs <- c(as.numeric(auc_base), as.numeric(auc_enh), as.numeric(auc_rf), as.numeric(auc_xgb))
names(aucs) <- c("base","enh","rf","xgb")
best <- names(aucs)[which.max(aucs)]
best
```

```{r importance}
if(best == "rf"){
  vip::vip(rf_model$finalModel, num_features = 15)
} else if(best == "xgb"){
  imp <- xgboost::xgb.importance(model = xgb_final)
  knitr::kable(head(imp, 15), caption = "XGBoost variable importance (top 15)")
} else if(best == "enh"){
  coefs <- broom::tidy(glm_enh) %>% arrange(desc(abs(estimate)))
  knitr::kable(head(coefs, 15), caption = "Top coefficients (enhanced logistic)")
} else {
  coefs <- broom::tidy(glm_base) %>% arrange(desc(abs(estimate)))
  knitr::kable(head(coefs, 15), caption = "Top coefficients (baseline logistic)")
}
```

---

# 10. Final scoring and export

```{r final-scoring}
# Bake full dataset using the prepared recipe
full_baked <- bake(prep_rec, new_data = data_full)

# Prepare final predictor set for chosen model
if(best %in% c("rf")){
  final_x <- full_baked %>% select(any_of(names(x_train)))
  final_probs <- predict(rf_model, newdata = final_x, type = "prob")[, "Yes"]
} else if(best == "xgb"){
  final_x <- full_baked %>% select(any_of(names(x_train)))
  final_matrix <- as.matrix(final_x %>% mutate(across(everything(), ~ as.numeric(as.character(.)))))
  dfinal <- xgboost::xgb.DMatrix(data = final_matrix)
  final_probs <- predict(xgb_final, newdata = dfinal)
} else if(best == "enh"){
  final_baked_enh <- bake(prep(rec_enh, training = train), new_data = data_full)
  final_probs <- predict(glm_enh, newdata = final_baked_enh, type = "response")
} else {
  final_baked_base <- bake(prep(rec_base, training = train), new_data = data_full)
  final_probs <- predict(glm_base, newdata = final_baked_base, type = "response")
}

data_full$Predicted_Risk <- final_probs

# Show sample
knitr::kable(head(data_full %>% select(PseudoID, Predicted_Risk, Attrition_Flag)), caption = "Sample final predictions")
```

---